---
title: "Data import"
---

<br>
<div class = "blue">
## Learning objectives

* Get comfortable importing different kinds of data
* Understand the concept of "tidy data"
</div>

<br>

```{r, include = FALSE}
library(tidyverse)
```

## Reading csv data

Data come in many forms, and we need to be able to load them in R. For our own 
use and with others who use R, there are R-specific data structures we can use, such as the `.rda` (`.RDATA`), or the `.rds` file-types. We'll talk about those in more detail later. However, we need to be able to work with more general data types too. Comma-separated value (`.csv`) tables are perhaps the most universal data structure. 

This [species dataset](https://gge-ucd.github.io/R-DAVIS/data/species.csv) provides genus and species information for animals caught
during a trapping survey. I downloaded it and put it in the `data` directory of my 
project. Click on the link and download the file.

We can read csv's with the `read.csv` function. The argument to `read.csv` is 
the location of the file, relative to your working directory. Since I saved the 
gapminder data to the `data` directory of my project, I can load it with this. 
Note the use of tab-completion to find the file and get it exactly right without 
typos.

```{r, eval = FALSE}

# download and read from your data folder
read.csv('data/species.csv')

```

Whoa! What just happened? R executed the function and printed the result, just 
like when you enter `log(1)`. How do you store an object to a variable?

```{r}

species <- read.csv('data/species.csv')

```

Now, a data.frame called `species` is in my Environment, and I can see that it 
is a data.frame with `r nrow(species)` rows and `r ncol(species)` columns. 

<div class = "blue">
### Challenge -- read csv data

The species data are available at [this link](data/species.csv).  
 
 - Right click on the link to "save file as..."
 - Save the .csv file in the `/data` directory of your project.
 - Read the data with the `read.csv` function and assign it to the variable `species`.  
 - Inspect the `data.frame` using whatever means you like. 
 - Use a `dplyr` verb to reorder the data frame in alphabetical order by genus
</div>

<br>
 
<div class = "blue">
### Advanced challenge

Suppose you get a .csv file from a colleague in Europe. Because they use `,` (comma) as a decimal separator, they use `;` (semi-colons) to separate fields.  
 - How can you read it into R? Feel free to use the web and/or R's helpfiles.  
</div>

<br>

### Using `stringsAsFactors=FALSE`

By default, when building or importing a data frame, the columns that contain 
characters (i.e., text) are coerced (=converted) into the factor data type. 
Depending on what you want to do with the data, you may want to keep these 
columns as character. To do so, `read.csv()` and `read.table()` have an argument 
called `stringsAsFactors` which can be set to `FALSE`.

<div class = "blue">
**Tip**: In most cases, it’s preferable to set `stringsAsFactors = FALSE` 
when importing your data, and converting as a factor only the columns that 
require this data type.
</div>
<br>

Compare the output of `str(surveys)` when setting `stringsAsFactors = TRUE` 
(default) and `stringsAsFactors = FALSE`:

```{r eval = FALSE, echo = TRUE}
surveys <- read.csv('data/combined.csv')

## Compare the difference between when the data are being read as
## `factor`, and when they are being read as `character`.
surveys <- read.csv("data/combined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/combined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)

```

## Reading messier data

Sometimes data can be a bit trickier to read because it isn't in tidy format.
If it is **close** to being tidy, we may be able add some more arguments to the
`read.csv()` call to to help R interpret how the file should be read. If there
are a few things that R will need help with, using the `read.table()` gives you
the most flexibility but you'll also have to always specify more information
about the file to help R interpret it.

Use [this link](data/wide_eg.csv) to download this "wide" dataset and inspect it with
a spreadsheet program. Why isn't it tidy?

Try using the `read.csv()` function on the file.

```{r, eval = FALSE}
read.csv("data/wide_eg.csv")
```

We need to tell R to skip 2 rows!

```{r, eval = FALSE}
read.csv("data/wide_eg.csv", skip = 2)
```

Or using `read.table()`...

```{r, eval = FALSE}
read.table("data/wide_eg.csv", header = TRUE, sep = ",", skip = 2)
```

## Reading directly from websites

We can use the `read.csv()` function to read files directly from a website if we
first call the `url()` function on the web address.

```{r wwwReadCsv, eval = FALSE}
read.csv(url("https://mikoontz.github.io/data-carpentry-week/data/wide_eg.csv"), skip = 2)
```

We can also use the `read_csv()` (note that's an underscore in the function
name, not a period) from the `readr` package which is part of the `tidyverse`.
In this case, we don't have to use the `url()` function.

```{r, eval = FALSE}
read_csv("https://mikoontz.github.io/data-carpentry-week/data/wide_eg.csv", skip = 2)
```

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new datasets to share
them with your collaborators or for archival.

Similar to the `read.csv()` function used for reading CSV files into R, there is
a `write.csv()` function that generates CSV files from data frames.

Before using `write.csv()`, we are going to create a new folder, `data_output`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data_output`
directory, so even if the files it contains are deleted, we can always
re-generate them.

 - Type in `write.` and hit TAB. Scroll down and take a look at the many options that exist for writing out data in R. 

The `tidyverse` has a similar set of functions that are part of the `readr` package. 

 - Type in `write_` and hit TAB. Use the F1 key on any of these options to read more about how to use it.

## Using `save` and `load`

We've seen how to save out specific data types in R. One additional option is you can save anything in your workspace or working "environment". You can save a single object, or multiple objects together. This is best done using the `.rda` file type. The **.Rda**ta file type is a great way to save objects from R that you have already tidied or modeled, and you want to maintain the structure and format of the data. Because these are native "R" file types, they can be read back in and will appear exactly as when you saved them. 

Let's assume we have the "surveys" object and the "counts" objects in our environment, and we want save them both in one file. We'll use `save()` and the `.rda` filetype.

```{r, eval=FALSE}
save(surveys, counts, file = "data_output/surveys_counts.rda")
```

To re-load these back into R, we use the `load` function. 

```{r loaddata, eval=F}

load("data_output/surveys_counts.rda")

```

### `.rds` vs. `.rda`

Why use one vs. the other? What do these file types provide that a simple csv doesn't provide? 

Short answer is they maintain not only the structure, but also the format and data types within your data sets. However it appears in your environment in R is exactly how it will be saved (with `save()`), and then read back in (with `load()` or `readRDS`. This can save time and code, once you have your data in a format/shape you are happy with. An added bonus is the format (`rda` and `rds`) are both highly compressed, and can save significant space on your hard drive. 

The main differences between the two:

 - `.rda` allows saving multiple objects together in one file, or one single object
 - `.rds` can only save file/object
 
For example, let's use the surveys dataset we loaded earlier and save a few `rda` and `rds` files.
 
```{r rda_vs_rds, eval=F, echo=T}

library(dplyr)

# load data
surveys <- read.csv('data/combined.csv') # the combined.csv is 3.1 MB in size

# filter to years > 2000 and only rodents
rodents2001 <- surveys %>% filter(year > 2000 & taxa == "Rodent")

# filter to only birds
birds <- surveys %>% filter(taxa=="Bird")

# RDA files: now we can save all these objects together:
save(surveys, rodents2001, birds, file = "data/example.rda") # this file is 245 kb (0.2 MB)

# RDS files: this can only be one file:
saveRDS(rodents2001, file = "data/rodents2001.rds")

# to load an rds file back in:
any_name_i_choose <- readRDS(file = "data/rodents2001.rds")

# but won't work with rda
nope <- readRDS(file = "data/example.rda")

# same with trying to assign a name to an rda file on load
doublenope <- load(file = "data/example.rda") # wait ...check out "doublenope"

doublenope # created a character vector of the objects...but didn't assign any to the name nope

# [1] "surveys"     "rodents2001" "birds"   
``` 

One extremely time-saving use of `.rds` files is when you've fit a large statistical model that gets saved as a model object. For example, the `brms` and `lme4` packages used for multi-level models will return large model objects that contain lots of information about the model. Some models can take hours, days, even **weeks** to finish fitting, so it can be useful to save a fully fitted model object as a `.rds` file. You can then load that model object back into R and work with it however you want.

## Excel & GoogleSheets

 - [`readxl`](http://readxl.tidyverse.org/) (Part of `tidyverse`)
 - Jenny Bryan's `googlesheets` or [`googledrive`](http://googledrive.tidyverse.org/) packages

## Other File Types

 - Using the `foreign` package
 - reading in `.dbf`, *Stata*, *SAS*, *SPSS*, `.shp`, `.netcdf`, `raster`, `.kml`, `gpx`, `xml`, `geojson`, `json`, etc....
 
## `rio` Package

One other data import/export package to check out is called `rio`. It uses two main functions: `import` and `export`, and it automatically detects the file extension you're using, then picks the fastest function out of a bunch of different specialized packages. It can streamline your data import/export and give you more consistent data frames when you're working with lots of different file types.

```{r}
#install.packages("rio")
library(rio)
export(mtcars, "data/mtcars.csv")
export(mtcars, "data/mtcars.rds")
export(mtcars, "data/mtcars.feather")
```

You can read more about `rio` on its [Github page](https://github.com/leeper/rio).

<br>

This lesson is adapted from the Software Carpentry: R for Reproducible 
Scientific Analysis [Vectors and Data Frames materials](http://data-lessons.github.io/gapminder-R/03-data-types-subsetting.html)
and the Data Carpentry: R for Data Analysis and Visualization of Ecological Data 
[Exporting Data materials](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html).