---
title: Vectors and data frames
minutes: 50
---

> ## Learning objectives
>
> - Understand vectors as fundamental R units
> - Understand and be able to use vectorized operations
> - Understand `data.frame` as common R object and familiar research data table.
>     - Columns = variables, rows = observations
>     - Columns = vectors
> - Be able to examine the structure and content of a data frame.
> - Be able to subset vectors with index and logical notation
>

## Intro to vectors

Vectors are a collection of observations of a single variable. They are the 
fundamental unit of R. In data analysis and statistics, we don't often work with 
individual numbers, but multiple observations. This is baked into R and helps  
make it powerful. Let's explore how to work with them.

### Creating vectors

We can also manually create new vectors. There are many ways to do this. 

We also use the `:` operator as a shortcut to generate a sequence of numbers 
starting from the number on the left of the `:` and going to the number on the 
right side of the `:`.

```{r colon}
3:10
```

The most flexible way is the concatenate function, `c()`. We can concatenate any
number of objects together, as long as they are the same type. Here's a vector
of characters.

```{r c}
fruit <- c("apples", "oranges", "lemons")
```

Let's store a vector of 6 odd numbers in a new variable. Remember that you must 
assign the result of an operation to a variable if you want to keep it around. 
Otherwise R will print out the result but then forget it.

```{r c_odds}
odds <- c(3, 5, 7)
```

### Vectorization

We can do math on a vector and the operation is performed on each element in
turn. We can reassign variables just as we did when they weren't vectors.

```{r vector_math}
odds + 1
odds <- odds - 2
```

Most functions that accept a single value can also accept a vector of values.

```{r log_vector}
exp(odds)
```

## Intro to loading external data 

We'll talk more about importing data later but, for now, use 
[this link](data/continents.RDA) to download a dataset that we'll use for the 
next section. Save it in the "data" folder you created earlier.

R has a variety functions to load specific kinds of data. Two R-specific 
data-types are `.RDA` and `.RDS`. `.RDA` files, like the one you just 
downloaded, are created with the `save()` function and accessed with the 
`load()` function. `load()` needs the location of the saved file, provided as 
character string file-path, starting with the working directory. If you are 
using RStudio's projects (which we recommend), this is made a bit easier because 
the location of your project (where the `.Rproj` file is located) is the 
default working directory. It is displayed at the top of your console pane in 
RStudio. File-paths should be provided relative to that location. So, to load 
the file we just saved:

```{r load data}
load('data/continents.RDA')
```

## Intro to data frames

```{r df, echo = FALSE}
continents
```

This is a `data.frame` -- the type of object most of us work with most of the 
time in R. In a `data.frame` each column represents a variable, and each row is 
an observation. This is the basic data organizational scheme -- one column per 
variable, one row per observation. You might recognize this form from from a 
statistics class or your own data analysis. 

#### Inspecting a data frame

Rather than pulling up the spreadsheet form of a data.frame, we'd like to use R 
to get more information about it. In this case, our `data.frame` is so small 
that we can print the whole thing and inspect it:

```{r}
continents
```

When we start working with more-realistic datasets, let alone big data, that 
won't work. We can get the first few rows of a `data.frame` with `head()`. 

```{r}
head(continents)
```

`str` provides richer information about a `data.frame` and each element in it. 
It is generally a good first-step inspection of an R object.

```{r}
str(continents)
```

We get some summary information on `continents`: it's type and dimensions, and 
we get some information on each variable in the `data.frame`. 

> #### Shoutout
>
> There is another function like `head()` and `str()` that provides information 
> on a `data.frame`: `summary()`  
> - Call the `summary` function on the `continents` data.  
> - What is the average (mean) change in tooth growth overall?
>

### Accessing a data frame 

We can extract individual columns of a data frame as vectors. 

We can extract a vector (ie a single variable) from a data frame with the `$` 
operator.

```{r extract}
continents$population
```

Note that you can use tab-completion to see what variables are available.

That just printed the six values of `population`. We are going to work with them 
some more, so extract them from the data.frame and store them to a new object, 
called `pop`:

```{r extract and assign}
pop <- continents$population
```

Now we have a new object in our environment: a numeric "vector" with 6 entries. 

Another core R concept is the idea that when you manipulate an object, the 
original object doesn't change. Note that the `continents` data.frame still has 
the dose variable. R didn't "take it out" of `continents`; instead, it made a 
copy of it and stored it to a variable called `pop`. They are now two 
separate things. If we make a change to one, it will not affect the other. 

#### Vectorization with data frames

We can now pair the concept of vectorization with operations on our data frame.
Here, we can find the logarithm of each continent's population:

```{r first vectorization}
log(x = pop, base = 10)
```

R went over each item in `pop` and took the base-10 logarithm. Some functions 
take a vector but rather than returning a result for each item, operate on all 
of them together. E.g. we can find the mean and standard deviation of 
populations:

```{r mean and sd}
mean(pop)
sd(pop)
```

Many functions will also operate element-wise over multiple vectors. E.g., to 
calculate the population density of each continent, we need to divide the 
population by the land area for each continent. We can do that with a single 
command:

```{r}
continents$population / continents$area_km2
```

Note that we didn't have to take those vectors out of the data.frame to use 
them. We can do vectorized operations right in the data.frame, which helps keep 
everything organized: recall that each row of a data.frame is a particular 
observation (in this case a continent), so we often want to do operations with 
each row.

Just like we can extract a column from a data.frame with `$`, we can make a new 
column:

```{r new column}
continents$pop_density <- continents$population / continents$area_km2
continents
```

### Subsetting

#### Subset by index

We can extract items from a vector by specifying which positions, or indices, we 
want. R's syntax for subsetting is square brackets (`[ ]`) at the end of an 
object containing the positions to return. So to get the third element out of 
our `pop` vector:

```{r subsetting1}
pop
pop[3]
```

To get the first three elements, we need to put a vector containing 1, 2, and 3 
inside the `[ ]`. We just learned how to make such a vector using the combine 
function.

```{r}
pop[c(1, 2, 3)]
```

Sometimes it will be more useful to provide the desired positions as a variable. 
Let's pull out the odd-positioned entries from `pop`:

```{r}
pop[odds]
```

We can also tell R which elements we *don't* want with a `-`. This returns each 
element in `pop` except the first one:

```{r}
pop[-1]
```

If we try to ask for an element that doesn't exist, R returns `NA`. `NA` is a 
special value in R that means "missing".

```{r}
pop[10]
```


> #### Challenge -- Create and subset a vector
>
> Similar to `c`, the `seq` function creates a vector: a **seq**uence of 
> numbers. 
>
> - Your first task is to create a sequence of all the multiples of three from 
> three to 300. Figure out how to do this. Some combination of playing with the 
> function in the console and reading its helpfile (`?seq`) should work. 
> Helpfiles are challenging at first, but it's important to learn how to find 
> the info you need in them. Hint: The arguments you need here are `from`, `to`, 
> and `by`.
>
> - Store your sequence in a variable.
> - Extract the 33rd entry from the sequence
> - **Bonus**: Create a vector with ten evenly-spaced numbers starting with one 
> and ending with one-million. What is the sum of the second and ninth entries 
> in this vector?
> - **Super-Bonus**: Returning to the multiples-of-three vector, what is the sum 
> of the numbers in positions that are not evenly divisible by three? That is, 
> the sum of the first, second, fourth, fifth, seventh, etc. entries.
>

#### Subset by logical

Note that the `continents` data frame contains two common kinds of variables: 
numbers (num) and strings (chr).

```{r}
str(continents)
```

A third important data type in R is logicals. You saw this when making logical 
comparisons like `1 > 0`. Logical data can only be `TRUE` or `FALSE` (or `NA` 
for missing; any variable type in R can have `NA`-missing values). 

Logical comparisons are vectorized like other things in R. Let's find the highly 
populated continents, those with more than one-billion people. One-billion is 
10^9, so we can write it as `1e9`.

```{r}
pop > 1e9
```

That went over each element in `pop` and compared it with 10. We say that R 
"recycled" 10 to compare it with each element in `pop`. 

In the same way that we subset by index before, we can subset by a logical 
vector. To find the values of `pop` that were greater than one-billion, we 
subset it like so:

```{r}
pop[pop > 1e9]
```

> #### Shoutout
>
> How could you extract the names of the continents with more than one-billion
> people?
>

Recall that in a data frame, each row constitutes a single observation. This 
makes it especially useful to test one column and use it to subset another -- we 
often want the entries of some variable where a condition on another variable is 
met. For example, to find the land-area of Africa, we can test the continent 
names for being "Africa" and extract the area where that is true. 

Here as elsewhere, it is often useful to build code from the inside out, i.e., 
write the logical test first, then go left and write what you want to subset 
with it.

```{r}
continents$area_km2[continents$continent == "Africa"]
```

> #### MCQ -- Subset and vectorize 
>
> What is the total land area of continents that have at least 10% of the 
> world's population? 
>
> - Use subsetting to get the areas you want
> - Use the `sum()` function to find the total land area
>
> a. 149428500
> b. 126700000
> c. 22728500
> d. 100
>

This lesson is adapted from the Software Carpentry: R for Reproducible 
Scientific Analysis [Vectors and Data Frames materials](http://data-lessons.github.io/gapminder-R/03-data-types-subsetting.html).



