---
title: "Coding Practice"
output:
  html_document:
    toc: false
---

The following material will be posted on Github via the Discussion board for the class *Team* (2019 is the **Hippogriffs**). Assignments are designed to reinforce the code/lessons covered that week, and provide students a chance to both practice coding and reviewing code. Assignments were completed via private repositories each student set up (via [Github Classrooms](https://classroom.github.com/classrooms)), and each week students will be randomly assigned another student for a code review. Apologies if there are links that don't work, they may be specifically setup for students enrolled in course.

<br>

## Assignments {.tabset .tabset-fade .tabset-pills}

### <small>Week2</small>

### <small>Week3</small>

### <small>Week4</small>

**Week 4 Assignment**

For you assignment this week, we will be practicing subsetting the same data frame we were working with in class, `surveys`. 

The objective of the assignment is for you to write a new, clean script, commit and push it to github, and then someone else will download and run that script on their computer, and file an issue in your repository with some comments about the script.

**Part I: Coding Practice**

1. Create a new script in your `r-davis-in-class-yourname` project named `week_4_assignment_ABC.R` but with your initials in place of ABC. Save it to your `scripts` folder.

2. Read `portal_data_joined.csv` into R using the `read.csv` command and create a dataframe named `surveys` (just like we did in class). 

3. Subset to just the first column and columns five through 8. Include only the first 400 rows. Save this as a dataframe called `surveys_subset`.

4. *CHALLENGE:* Select all rows that have a `hindfoot_length` greater than 32, save these in a new data.frame named `surveys_long_feet`, then plot its `hindfoot_length` values as a histogram using the `hist` function.

5. Change the column `hindfoot_lengths` into a character vector.

6. Plot the `hindfoot_lengths` in a histogram (if this doesn't work, just leave it, and think about it during Part II of the assignment, *wink wink*).

7. Commit & Push the R script to GitHub - check your repository on GitHub and make sure your script is there.

**Part II: Code Review**

Find your code partner’s repository on GitHub, click on the scripts folder, and find the script named `w3_assignment_ABC.R`

Click on the script, then find the RAW button on the right hand side of the screen. Click on RAW.

You should see a plain text version of the script. Right click anywhere on the page, and select “Save As”, save the script to your  scripts folder in your repository. Make sure to change the file extension to `.R` (instead of `.txt`).

Now, if you open up your RProject where you saved the script, you should see it, and better yet, you should have the same path/folder structure that is used in the script (we hope!).

Try to run the script line by line. Where does it break? Can you fix it?
File an Issue in your code partner’s repository. Write up a few sentences describing a few good things about the script (what worked! Was it clearly commented?) and what didn’t (where did it break for you, and why? How might it be fixed?). Remember to be constructive and positive…we’re all learning!

Any questions? comments? Issues? Remember you can always tag us using individual names when you file an issue in the [Discussion repository](https://github.com/gge-ucd/Discussion/issues): @marthawohlfeil, @mcmaurer @ryanpeek, etc.

Thanks!

### <small>Week5</small>

**WEEK 5 ASSIGNMENT**

Alright folks, we've ventured into the `tidyverse`! We're going to test out your `dplyr` skills, pipe (`%>%`) some data, and make a `tibble` or two.

**Part I: Enter the Tidyverse**

1) Create a new script in your `r-davis-in-class-yourname` project, and name it `w_5_assignment_ABC.R`, with your initials in place of `ABC`. Save it in your `scripts` folder.

2) Read `portal_data_joined.csv` into R using the `tidyverse`'s command called `read_csv()`, and assign it to an object named `surveys`.

3) Using `tidyverse` functions and pipes, subset to keep all the rows where weight is between 30 and 60, then print the first few (maybe... 6?) rows of the resulting tibble.

4) Make a tibble that shows the `max` (hint hint) weight for each species+sex combination, and name it `biggest_critters`. Use the `arrange` function to look at the biggest and smallest critters in the tibble (use `?`, tab-complete, or Google if you need help with `arrange`).

5) Try to figure out where the `NA` weights are concentrated in the data- is there a particular species, taxa, plot, or whatever, where there are lots of `NA` values? There isn't necessarily a right or wrong answer here, but manipulate `surveys` a few different ways to explore this. Maybe use `tally` and `arrange` here.

6) Take surveys, remove the rows where weight is `NA` and add a column that contains the average weight of each species+sex combination. Then get rid of all the columns except for species, sex, weight, and your new average weight column. Save this tibble as `surveys_avg_weight`. The resulting tibble should have 32,283 rows. 

7) *Challenge*: Take `surveys_avg_weight` and add a new column called `above_average` that contains logical values stating whether or not a row's weight is above average for its species+sex combination (recall the new column we made for this tibble).

8) *Extra Challenge*: Figure out what the `scale` function does, and add a column to `surveys` that has the scaled weight, by species. Then sort by this column and look at the relative biggest and smallest individuals. Do any of them stand out as particularly big or small?

**Part II: Code Review**

Find your code partner’s repository on GitHub, click on the scripts folder, and find the script named `w5_assignment_ABC.R`

Click on the script, then find the RAW button on the right hand side of the screen. Click on RAW.

You should see a plain text version of the script. Right click anywhere on the page, and select “Save As”, save the script to your  scripts folder in your repository. Make sure to change the file extension to `.R` (instead of `.txt`).

If you can't save your partner's file in this way, you can copy the text from the RAW view, then create a new `.R` file in your RProject, and paste the text there.

Now, if you open up your RProject where you saved the script, you should see it, and better yet, you should have the same path/folder structure that is used in the script (we hope!).

Try to run your partner's script, stepping through each question, and compare your methods and answers to theirs. Did they use the same steps in the same order? Do their answers look like yours? If not, try to figure out why!

Any questions? comments? Issues? Remember you can always tag us using individual names when you file an issue in the [Discussion repository](https://github.com/gge-ucd/Discussion/issues): @marthawohlfeil, @mcmaurer @ryanpeek, etc.

Thanks!

### <small>Week6</small>


This week's assignment builds on the `ggplot` skills that we learned during Week 06. Just like last week, there are two parts: the exercsise and a code review. Please pick a partner for the code review as you have been doing in past weeks. If you do not have a partner, please email Martha at mwohlfeil@ucdavis.edu. 

#### **Part 1**

This exercise asks you to **modify existing code to complete a new task.**

We will work in the same repository that you've been working in for the last couple of weeks: r-davis-in-class-YourName.

For this homework, we are going to be working with a new dataset called `gapminder`. Download the data using the code below. Remember, this code is looking for a folder called `data` to put the .csv in, so make sure you have a folder named `data`, or modify the code to the correct folder name. 

```{r}
library(tidyverse)
gapminder <- read_csv("https://gge-ucd.github.io/R-DAVIS/data/gapminder.csv")

```


#### *1A*.

Modify the following code to make a figure that shows how life expectancy has changed over time:

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
geom_point()
```


#### *1B*.

Look at the following code. What do you think the `scale_x_log10()` line is doing? What do you think the `geom_smooth()` line is doing? 

**Hint:** There's no cost to tinkering! Try some code out and see what happens with or without particular elements.

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent), size = .25) + 
    scale_x_log10() +
    geom_smooth(method = 'lm', color = 'black', linetype = 'dashed') +
    theme_bw()
```


#### *1C*. (Challenge!)

Modify the above code to size the points in proportion to the population of the county. 
**Hint:** Are you translating data to a visual feature of the plot?

#### Save this code in your scripts folder and name is `w6_assignment_ABC.R` with your initials. 

#### **Part II: Code Review**

 - Find your code partner’s repository on github, click on the scripts folder, and find the script named: `w6_assignment_ABC.R`
 - Click on the script, then find the `RAW` button on the right hand side of the screen. Click on `RAW`.
 - You should see a plain text version of the script. Right click anywhere on the page, and select “**Save As**”, save the script to your `scripts` folder in your repository. Make sure to change the file extension to `.R` (instead of `.txt`).
 - Now, if you open up your RProject where you saved the script, you should see it, and better yet, you should have the same path/folder structure that is used in the script (*we hope!*).
 - Try to run the script line by line. Where does it break? Can you fix it?
 - **File an Issue** in your code partner’s repository. Write up a few sentences describing a few good things about the script (*what worked! Was it clearly commented?*) and what didn’t (*where did it break for you, and why? How might it be fixed?*). Remember to be constructive and positive...we're all learning!

### <small>Week7</small>

### <small>Week8</small>

### <small>Week9</small>



