---
title: "Coding Practice"
output:
  html_document:
    toc: false
---

The following material will be posted on Github via the Discussion board for the class *Team* (2020 is the **Dragons**). Assignments are designed to reinforce the code/lessons covered that week, and provide students a chance to both practice coding and reviewing code. Assignments were completed via private repositories each student set up (via [Github Classrooms](https://classroom.github.com/classrooms)), and each week students will be randomly assigned another student for a code review. Apologies if there are links that don't work, they may be specifically setup for students enrolled in course.

<br>

## Assignments {.tabset .tabset-fade .tabset-pills}

### <small>Week2</small>

### <small>Week3</small>

### <small>Week4</small>

Any questions? comments? Issues? Remember you can always tag us using individual names when you file an issue in the [Discussion repository](https://github.com/gge-ucd/Discussion/issues): @marthawohlfeil, @mcmaurer @ryanpeek, etc.


### <small>Week5</small>

### <small>Week6</small>

### <small>Week7</small>

### <small>Week8</small>
details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}

```
</details>

### <small>Week9</small>
### Answers:
<details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}
            
```
</details>

