---
title: "Coding Practice"
output:
  html_document:
    toc: false
---

The following material will be posted on Github via the Discussion board for the class *Team* (2020 is the **Dragons**). Assignments are designed to reinforce the code/lessons covered that week, and provide students a chance to both practice coding and reviewing code. Assignments were completed via private repositories each student set up (via [Github Classrooms](https://classroom.github.com/classrooms)), and each week students will be randomly assigned another student for a code review. Apologies if there are links that don't work, they may be specifically setup for students enrolled in course.

<br>

## Assignments {.tabset .tabset-fade .tabset-pills}

### <small>Week2</small>

Today we talked quite a bit about project management, directory structure, and file paths. Understanding how your computer thinks about file organization can be tricky, but it's crucial to any computational task and good project management. We'll have you do a little exercise to help sketch out the relationships between files and folders in your computer.

1) Grab a pen/pencil and paper and draw out some sort of diagram showing the structure of your "in-class-project" directory. Include the sub-folders and some example files. Somewhere in this diagram, denote what git views as your "repository" and what RStudio views as your "R Project" (Hint: this might involve circling portions of your diagram). Feel free to add notes about the purposes of files/folders you included, or even add new files and folders you think could aid in your project organization. **You're going to be taking a picture of this diagram/uploading it, so if you'd prefer to make it digitally, like in PowerPoint or Lucidchart, go right ahead**

2) If you drew your diagram on paper, take a picture of it and email it to yourself. If you made it digitally, save it as a .png or .jpg or take a screenshot of it. Get the photo of your diagram into your "in-class-project" folder, and put it into an "images" folder there. You should end up with a file that's something like "in-class-project-yourname/images/your_diagram.jpg". This will be practice for taking external files, moving them into a project, and using them within the project.

3) Get on the good ole Google and look up how to add a local image to a Markdown file. Your README.md is written in Markdown, and we're going to add your diagram photo to it. This should involve just one line of new code- the trickiest part is referring to the correct location of your photo relative to the README file (Hint: your "link" to the file won't be a website link). Once you've made this change, commit and push your changes to GitHub. This photo should now appear in the README when you open your repository on GitHub.

### <small>Week3</small>

### <small>Week4</small>

Any questions? comments? Issues? Remember you can always tag us using individual names when you file an issue in the [Discussion repository](https://github.com/gge-ucd/Discussion/issues): @marthawohlfeil, @mcmaurer @ryanpeek, etc.


### <small>Week5</small>

### <small>Week6</small>

### <small>Week7</small>

### <small>Week8</small>
details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}

```
</details>

### <small>Week9</small>
Answers:
<details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}
            
```
</details>

