---
title: "Coding Practice"
output:
  html_document:
    toc: false
---

The following material will be posted on Github via the Discussion board for the class *Team* (2020 is the **Dragons**). Assignments are designed to reinforce the code/lessons covered that week, and provide students a chance to both practice coding and reviewing code. Assignments were completed via private repositories each student set up (via [Github Classrooms](https://classroom.github.com/classrooms)), and each week students will be randomly assigned another student for a code review. Apologies if there are links that don't work, they may be specifically setup for students enrolled in course.

<br>

## Assignments {.tabset .tabset-fade .tabset-pills}

### <small>Week2</small>

Today we talked quite a bit about project management, directory structure, and file paths. Understanding how your computer thinks about file organization can be tricky, but it's crucial to any computational task and good project management. We'll have you do a little exercise to help sketch out the relationships between files and folders in your computer.

1) Grab a pen/pencil and paper and draw out some sort of diagram showing the structure of your "in-class-project" directory. Include the sub-folders and some example files. Somewhere in this diagram, denote what git views as your "repository" and what RStudio views as your "R Project" (Hint: this might involve circling portions of your diagram). Feel free to add notes about the purposes of files/folders you included, or even add new files and folders you think could aid in your project organization. **You're going to be taking a picture of this diagram/uploading it, so if you'd prefer to make it digitally, like in PowerPoint or Lucidchart, go right ahead**

2) If you drew your diagram on paper, take a picture of it and email it to yourself. If you made it digitally, save it as a .png or .jpg or take a screenshot of it. Get the photo of your diagram into your "in-class-project" folder, and put it into an "images" folder there. You should end up with a file that's something like "in-class-project-yourname/images/your_diagram.jpg". This will be practice for taking external files, moving them into a project, and using them within the project.

3) Get on the good ole Google and look up how to add a local image to a Markdown file. Your README.md is written in Markdown, and we're going to add your diagram photo to it. This should involve just one line of new code- the trickiest part is referring to the correct location of your photo relative to the README file (Hint: your "link" to the file won't be a website link). Once you've made this change, commit and push your changes to GitHub. This photo should now appear in the README when you open your repository on GitHub.

### <small>Week3</small>

In Week 3's homework we are going to practice subsetting and manipulating vectors. Fun! 

First, open your r-davis-in-class-project-YourName and `pull`! Remember, we always want to start working on a github project by pulling, even if we are sure nothing has changed (believe me, this small step will save you lots of headaches).

Second, open a new script in your r-davis-in-class-project-YourName and save it to your `scripts` folder. Call this new script `week_3_homework`. 

Copy and paste the chunck of code below into your new `week_3_homework` script and run it. This chunck of code will create the vector you will use in your homework today. Check in your environment to see what it looks like. What do you think each line of code is doing? 

```{r}
set.seed(15)
hw3 <- runif(50, 4, 50)
hw3 <- replace(hw3, c(4,12,22,27), NA)
hw3
```

*PROBLEM 1:*

Take you `hw3` vector and removed all the NAs then select all the numbers between 14 and 38 inclusive, call this vector `prob1`. 

*PROBLEM 2:* 

Multiply each number in the `prob1` vector by 3 to create a new vector called `times3`. Then add 10 to each number in your `times3` vector to create a new vector called `plus10`. 

*PROBLEM 3:*

Select every other number in your `plus10` vector by selecting the first number, not the second, the third, not the fourth, etc. If you've worked through these three problems in order, you should now have a vector that is 12 numbers long that looks **exactly** like this one:

```{r, echo = F}
prob1 <- hw3[!is.na(hw3)] #removing the NAs

prob1 <- prob1[prob1 >14 & prob1 < 38] #only selecting numbers between 14 and 38

times3 <- prob1 * 3 #multiplying by 3

plus10 <- times3 + 10 #adding 10 to the whole vector 

final <- plus10[c(TRUE, FALSE)] #selecting every other number
```

```{r}
final
```

Finally, save your script and push all your changes to your github account! 

<details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=T}

prob1 <- hw3[!is.na(hw3)] #removing the NAs

prob1 <- prob1[prob1 >14 & prob1 < 38] #only selecting numbers between 14 and 38

times3 <- prob1 * 3 #multiplying by 3

plus10 <- times3 + 10 #adding 10 to the whole vector 

final <- plus10[c(TRUE, FALSE)] #selecting every other number using logical subsetting

```
</details>


### <small>Week4</small>

Homework this week will be playing with the `surveys` data we worked on in class. First things first, open your r-davis-in-class-project and PULL! Then create a new script in your `scripts` folder called `homework_week4.R`. 

We are going to ask you to use the two different types of subsetting techniques we learned about: Base R and `Tidyverse`. It's important to understand both, although in your own work you will probably be using just one. 

**PROBLEM 1:**

**First, Base R** (DON'T use any tidyverse functions for this part!)

Load your `survey` data frame with read.csv. Create a new data frame called `surveys_base` with only the species_id, the weight, and the plot_type columns. Have this data frame only be the first 60 rows. Convert both species_id and plot_type to characters. Remove all rows with weight == NA.

CHALLENGE!
Create a second data frame called `challenge_base` that only consists of individuals from your `surveys_base` data frame with weights greater than 150g.  

<details>
<summary>**DO NOT OPEN** until you are ready to see the answers for the Base R portion of the homework</summary>
```{r, eval=T}
#PROBLEM 1

surveys <- read.csv("data/portal_data_joined.csv") #reading the data in

colnames(surveys) #a list of the column names 

surveys_new <- surveys[1:60, c(6, 9, 13)] #selecting rows 1:60 and just columns 6, 9 and 13

surveys_new$species_id <- as.character(surveys_new$species_id) #converting factor data to character

surveys_new$plot_type <- as.character(surveys_new$plot_type) #converting factor data to character 

surveys_new <- surveys_new[complete.cases(surveys_new), ] #selecting only the ROWS that have complete cases (no NAs) **Notice the comma was needed for this to work**

#CHALLENGE
challenge_base <- surveys_new[(surveys_new[, 2]>150),] #selecting just the weights (column 2) that are greater than 150

```
</details>

<br> 

**PROBLEM 2:**

**Now, `tidyverse`!** For this second problem, you can use the tidyverse commands we learned in class (although you may need a base function or two). 

Load your `survey` data frame with `read_csv` and call it `surveys2`. Create a new data frame called `surveys_tidy` with only the species_id, hindfoot_length, and the year columns.  Remove all rows with hindfoot_length == NA. Next, select only the last 100 rows. *Remember, species_id should already be a character column, so there is no need to convert*

CHALLENGE!
Create a second data frame called `challenge_tidy` that only consists of the species "RM" in the year 1997. 

<details>
<summary>**DO NOT OPEN** until you are ready to see the answers for the Tidyverse portion of the homework</summary>
```{r, eval = T}
#PROBLEM 2

library(tidyverse) #load tidyverse 

surveys2 <- read_csv("data/portal_data_joined.csv") #read in data with tidyverse command

surveys_tidy <- surveys2 %>% 
  select(species_id, hindfoot_length, year) %>% 
  filter(!is.na(hindfoot_length)) #removing all the NAs from hindfoot_length

surveys_tidy <- surveys_tidy[-c(1:31338), ] #selecting the LAST 100 rows, data frame was 31438 rows long initially

#CALLENGE 
challenge_tidy <- surveys_tidy %>% 
  filter(year == 1997) %>% 
  filter(species_id == "RM")

```
</details>

**Finally, save your script and PUSH the changes to github.** 


### <small>Week5</small>

1. Create a tibble named `surveys` from the portal_data_joined.csv file.

2. Subset `surveys` to keep rows with weight between 30 and 60, and print out the first 6 rows.

3. Create a new tibble showing the maximum weight for each species + sex combination and name it `biggest_critters`. Sort the tibble to take a look at the biggest and smallest species + sex combinations. HINT: it's easier to calculate max if there are no NAs in the dataframe...

4. Try to figure out where the NA weights are concentrated in the data- is there a particular species, taxa, plot, or whatever, where there are lots of NA values? There isn’t necessarily a right or wrong answer here, but manipulate surveys a few different ways to explore this. Maybe use `tally` and `arrange` here.

5. Take `surveys`, remove the rows where weight is NA and add a column that contains the average weight of each species+sex combination to the **full** `surveys` dataframe. Then get rid of all the columns except for species, sex, weight, and your new average weight column. Save this tibble as `surveys_avg_weight`.

6. Take `surveys_avg_weight` and add a new column called `above_average` that contains logical values stating whether or not a row’s weight is above average for its species+sex combination (recall the new column we made for this tibble).

7. Manipulate `surveys` to create a dataframe with a column for genus and a column named after every plot type, with each of these columns containing the mean hindfoot length of animals in that plot type and genus. So every row has a genus and then a mean hindfoot length value for every plot type. The dataframe should be sorted by values in the Control plot type column. This question will involve quite a few of the functions you've used so far, and it may be useful to sketch out the steps to get to the final result.

<details>
<summary>**DO NOT OPEN** until you are ready to see the answers for the Tidyverse portion of the homework</summary>
```{r, eval = T}
library(tidyverse)
#1
surveys <- read_csv("data/portal_data_joined.csv")

#2
surveys %>% 
  filter(weight > 30 & weight < 60)

#3
biggest_critters <- surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id, sex) %>% 
  summarise(max_weight = max(weight))

biggest_critters %>% arrange(max_weight)

biggest_critters %>% arrange(desc(max_weight))

#4
surveys %>% 
  filter(is.na(weight)) %>% 
  group_by(species) %>% 
  tally() %>% 
  arrange(desc(n))

surveys %>% 
  filter(is.na(weight)) %>% 
  group_by(plot_id) %>% 
  tally() %>% 
  arrange(desc(n))

surveys %>% 
  filter(is.na(weight)) %>% 
  group_by(year) %>% 
  tally() %>% 
  arrange(desc(n))

#5
surveys_avg_weight <- surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id, sex) %>% 
  mutate(avg_weight = mean(weight)) %>% 
  select(species_id, sex, weight, avg_weight)

surveys_avg_weight

#6
surveys_avg_weight <- surveys_avg_weight %>% 
  mutate(above_average = weight > avg_weight)

surveys_avg_weight

#7
surveys %>% 
  filter(!is.na(hindfoot_length)) %>% 
  group_by(genus, plot_type) %>% 
  summarise(mean_hindfoot = mean(hindfoot_length)) %>% 
  pivot_wider(names_from = plot_type, values_from = mean_hindfoot) %>% 
  arrange(Control)

```
</details>

### <small>Week6</small>

For our week six homework, we are going to be practicing the skills we learned with ggplot during class. You will be happy to know that we are going to be using a brand new data set called `gapminder`. This data set is looking at statistics for a few different counties including population, GDP per capita, and life expentancy. Download the data using the code below. Remember, this code is looking for a folder called `data` to put the .csv in, so make sure you have a folder named `data`, or modify the code to the correct folder name. 

```{r, warnings = F}
library(tidyverse)

gapminder <- read_csv("https://gge-ucd.github.io/R-DAVIS/data/gapminder.csv") #ONLY change the "data" part of this path if necessary

```


#### *Problem 1*.

First calculates mean life expectancy on each continent. Then create a plot that shows how life expectancy has changed over time in each continent. Try to do this all in one step using pipes! (aka, try not to create intermediate dataframes)

#### *Problem 2*.

Look at the following code and answer the following questions. What do you think the `scale_x_log10()` line of code is achieving? What about the `geom_smooth()` line of code? 

*Challenge!* Modify the above code to size the points in proportion to the population of the country. 
**Hint:** Are you translating data to a visual feature of the plot?

**Hint:** There's no cost to tinkering! Try some code out and see what happens with or without particular elements.

```{r}


ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent), size = .25) + 
    scale_x_log10() +
    geom_smooth(method = 'lm', color = 'black', linetype = 'dashed') +
    theme_bw()

```


#### *Problem 3*

Create a boxplot that shows the life expectency for Brazil, China, El Salvador, Niger, and the United States, with the data points in the backgroud using geom_jitter. Label the X and Y axis with "Country" and "Life Expectancy" and title the plot "Life Expectancy of Five Countries".

<details>
<summary>**DO NOT OPEN** until you are ready to see the answers! </summary>
```{r, eval = T}

library(tidyverse)

#PROBLEM 1:

gapminder %>%
  group_by(continent, year) %>% 
  summarize(mean_lifeExp = mean(lifeExp)) %>% #calculating the mean life expectancy for each continent and year
  ggplot()+
  geom_point(aes(x = year, y = mean_lifeExp, color = continent))+ #scatter plot
  geom_line(aes(x = year, y = mean_lifeExp, color = continent)) #line plot

#there are other ways to represent this data and answer this question. Try a facet wrap! Play around with themes and ggplotly!  



#PROBLEM 2:

#challenge answer
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent, size = pop)) + 
    scale_x_log10() +
    geom_smooth(method = 'lm', color = 'black', linetype = 'dashed') +
    theme_bw()


#PROBLEM 3: 

countries <- c("Brazil", "China", "El Salvador", "Niger", "United States") #create a vector with just the countries we are interested in

gapminder %>% 
  filter(country %in% countries) %>% 
  ggplot(aes(x = country, y = lifeExp))+
  geom_boxplot() +
  geom_jitter(alpha = 0.3, color = "blue")+
  theme_minimal() +
  ggtitle("Life Expectancy of Five Countries") + #title the figure
  theme(plot.title = element_text(hjust = 0.5)) + #centered the plot title
  xlab("Country") + ylab("Life Expectancy") #how to change axis names


```
</details>


### <small>Week7</small>

For week 7, we're going to be working on 2 critical `ggplot` skills: recreating a graph from a dataset and **googling stuff**.

Our goal will be to make this final graph:

```{r, eval=T, echo=F, warning=F, message=F, }
library(tidyverse)

gapminder <- read_csv("data/gapminder.csv")

pg <- gapminder %>% 
  select(country, year, pop, continent) %>% 
  filter(year > 2000) %>% 
  pivot_wider(names_from = year, values_from = pop) %>% 
  mutate(pop_growth_0207 = (`2007` - `2002`) / 5)



pg %>% 
  filter(continent %in% c("Africa", "Americas", "Asia", "Europe")) %>% 
  ggplot(aes(x = reorder(country, pop_growth_0207), y = pop_growth_0207)) +
  geom_col(aes(fill = continent)) +
  facet_wrap(vars(continent), scales = "free") +
  theme_bw() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none") +
  xlab("Country") +
  ylab("Yearly Pop Growth Between 2002 and 2007")
```


<details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}

library(tidyverse)

gapminder <- read_csv("data/gapminder.csv")

pg <- gapminder %>% 
  select(country, year, pop, continent) %>% 
  filter(year > 2000) %>% 
  pivot_wider(names_from = year, values_from = pop) %>% 
  mutate(pop_growth_0207 = (`2007` - `2002`) / 5)



pg %>% 
  filter(continent %in% c("Africa", "Americas", "Asia", "Europe")) %>% 
  ggplot(aes(x = reorder(country, pop_growth_0207), y = pop_growth_0207)) +
  geom_col(aes(fill = continent)) +
  facet_wrap(vars(continent), scales = "free") +
  theme_bw() +
  scale_fill_viridis_d() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "none") +
  xlab("Country") +
  ylab("Yearly Pop Growth Between 2002 and 2007")
```
</details>

### <small>Week8</small>
details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}

```
</details>

### <small>Week9</small>
Answers:
<details>
<summary>**DO NOT OPEN** until you are ready to see the answers</summary>
```{r, eval=FALSE}
            
```
</details>

