---
title: Data visualization
minutes: 105 (1.75 hrs)
---

```{r, include=FALSE}
# Load the gapminder package so the gapminder data frame is available
library(gapminder)
# Create a results/ directory to match learners' projects, in which to save 
# plots:
dir.create('results')
```

> ## Learning objectives
>
> * To be able to use ggplot2 to generate publication quality graphics
> * To understand the basics of the grammar of graphics:
>   - The aesthetics layer
>   - The geometry layer(s)
>   - Adding statistics
>   - Transforming scales
>   - Modifying geom aesthetics by color/shape/size/linetype
>   - Faceting
>

Plotting is essential to understanding your data and the relationships in it. 
You should be making plots throughout the analysis and modeling process. There 
are three main plotting systems in R, the base plotting system, the lattice 
package, and the ggplot2 package.

We will learn `ggplot()`, because it is effective and efficient, makes 
publication-quality graphics, and is theoretically grounded in the grammar of 
graphics-- the idea that any plot can be expressed from the same set of 
components: a **data** set, a **coordinate system**, and a set of **geoms**--the 
visual representation of data points.

The `ggplot2` package is automatically installed with the `tidyverse` package,
so it will be loaded when we call the `library()` function on `tidyverse`.

```{r}
library(tidyverse)
```

### First plots

The key to understanding ggplot is thinking about a figure in layers: 
just like you might do in an image editing program like Photoshop, Illustrator, 
or Inkscape.

Let's start off with an example, longevity as a function of gdp:

```{r lifeExp-vs-gdpPercap-scatter, message=FALSE}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

The first thing we do is call the `ggplot` function. This function tells ggplot 
what dataset we're using (gapminder) and how to map variables in gapminder to 
the plotting "canvas" (gdp to the x-axis, life expectancy to the y-axis). 
That **mapping** of variables to plot elements is done through the `aes` 
function. Arguments to `aes` will be variable names from the `data.frame` passed 
to `ggplot`. If you need a reminder of what's in our data.frame, use 
`str(gapminder)`.

By itself, the call to `ggplot` isn't enough to draw a figure. This sets up the 
canvas correctly, but it doesn't draw anything on it.

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp))
```

Now we need to tell `ggplot` how to draw the data. For that, we add another 
layer: a **geom**. In our example, we used `geom_point`, which tells `ggplot` we 
want to draw a point for each row in the dataset. 

Note that we tell R that multiple lines belong to `ggplot` by including a `+` at 
the end of each line except the last.

```{r lifeExp-vs-gdpPercap-scatter2}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

> #### Challenge -- Changing a variable
>
> Modify the following code to make a figure that shows how life expectancy has
> changed over time:
>
> ```{r, eval=FALSE}
> ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
> geom_point()
> ```
>
> Hint: Inspect the data with `head(gapminder)` or `str(gapminder)` to find
> the name of the variable representing time. It should go on the x-axis.
>
>
> **Bonus**: So far, we have only mapped variables to the **x** and **y** 
> locations of each point.  Another *aesthetic* property we can modify is the 
> point `color`. Modify the code from the previous challenge to **color** the 
> points by the "continent" column. What trends do you see in the data? Are they 
> what you expected?
>

### Layers

Using a scatterplot probably isn't the best for visualizing change over time.
Let's add another layer to the plot -- lines connecting the points representing
each country:

```{r lifeExp-line-bad}
ggplot(gapminder, aes(x = year, y = lifeExp, color = continent)) +
    geom_point() +
    geom_line()
```

Hmm, that's not quite right. What has happened there? How many lines are there?

The lines are grouped by continent. We wanted one line per country, but we got 
one line per continent because we mapped continent to color and `ggplot` assumes 
we want to group by the coloring variable unless we tell it otherwise. To get 
one line per country we tell ggplot explicitly to group by country. Just like 
the other aesthetic mappings, that goes in the `aes()` function:

```{r lifeExp-line}
ggplot(gapminder, 
       aes(x = year, y = lifeExp, color = continent, group = country)) +
    geom_point() +
    geom_line()
```

Note that aesthetic mappings in the top-line `ggplot` call apply to all layers,
but you can also map variables just within a single layer. For example, if we
want colored lines but black points, we can move the mapping of 
continent-to-color from the `ggplot` line to the `geom_line`, so that it no 
longer applies to `geom_point`.

```{r lifeExp-line-2}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country)) +
    geom_point() +
    geom_line(aes(color = continent))
```

Note that each layer is drawn on top of the previous layer. If we want the 
points on top of the lines, we can move `geom_point` to after `geom_line`:

```{r lifeExp-layer-example-1}
ggplot(gapminder, aes(x = year, y=lifeExp, by=country)) +
    geom_line(aes(color = continent)) + 
    geom_point()
```

### Continuous vs. discrete variables

The plots we have done so far have plotted numeric variables to both axes. What 
if we want to plot a numeric variable against a categorical variable, such as 
continent. 

**Shoutout Q: What kind of plot can we use to plot a continuous variable **
**versus a categorical variable, say life expectency by continent?**

> NB: A barchart (or, less likely, a dotchart) is suboptimal because it requires 
summarizing the data to the mean or median or whatever -- we'd rather preserve 
the information present in the full distribution.

One possibility is a boxplot. In a boxplot, the distribution of values for each 
category is displayed, with the central line at the median value, the box 
representing the inner-50%, etc.

```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
    geom_boxplot()
```


> #### Challenge -- A new geom
>
> Violin plots are an alternative to boxplots that present the entire 
> distribution of the continuous variable. 
> 
> - Can you change the last plot from a boxplot to a violin plot?
> - Can you color each violin by its continent? 
>       - Tip: Try out the `fill` aesthetic.
>
> **Bonus**: Suppose you want to examine the distribution of life expectencies 
> across all countries together; that is, you want a single plot of showing the 
> distribution of all life-expectancy values. There are a few geom's that can do 
> this: Can you make one?


### Transformations

The grammar of graphics makes it easy to transform the scales on which the data 
are presented. To demonstrate we'll go back to our first example:

```{r lifeExp-vs-gdpPercap-scatter3, message=FALSE}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() 
```

Currently it's hard to see the relationship between the points due to some 
strong outliers in GDP per capita. We can change the scale of units on the 
x-axis using the *scale* functions. These control the mapping between the data 
values and visual values of an aesthetic.

The `log10` function applies a transformation to the values of the `gdpPercap`
column before rendering them on the plot, so that each multiple of 10 now only
corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the x-axis, a value of 10,000 corresponds to 4 on the 
x-axis and so on. This makes it easier to visualize the spread of data on the
x-axis.

```{r axis-scale}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() + 
    scale_x_log10()
```



### Adding models

We can fit a simple, linear relationship to the data by adding another layer,
`geom_smooth`, which adds a summary of a fit model to the plot. `method = 'lm'` 
tells `geom_smooth` to fit a **l**inear **m**odel. We could also fit other, more 
flexible smoothers by specifying different methods in `geom_smooth`.

```{r lm-fit}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point() + 
    scale_x_log10() +
    geom_smooth(method = 'lm')
```

> #### Challenge -- The power of money, by continent
>
> - Modify the last plot to color points by continent and get a separate 
fit-line for each continent. 
> - Tip: This can be done in one place, with fewer than 20 characters.
> - **Bonus**: Color the points by continent, but have only one line, fit to all 
> of the data.

### Adjust appearances

Some people like the gray `ggplot` background, others call it "chart junk" -- 
distracting, uninformative ink. You can change the overall appearance of a plot 
by setting a new theme. E.g. `theme_bw` is a bit cleaner:

```{r themes}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent)) + 
    scale_x_log10() +
    geom_smooth(method = 'lm') +
    theme_bw()
```

There are a variety of themes in a separate package called `ggthemes`, e.g. we 
can mimic The Economist [or many others](https://github.com/jrnold/ggthemes). Of 
course before we use the `ggthemes` package, we have to install it, which you 
can do with `install.packages('ggthemes')`.

```{r ggthemes}
library(ggthemes)
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent)) + 
    scale_x_log10() +
    geom_smooth(method = 'lm') +
    theme_economist()
```


#### Changing fixed aspects of plots

Of course, rather than changing the theme of an entire plot, we often want to 
changes some specific aspect of it. Let's make our linear model line black 
instead of blue and dashed instead of solid, and let's make our points smaller. 

Note that when aesthetics are mapped from a variable, they go in `aes()`, but to 
set them to a fixed value, they go outside `aes()`.


```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
    geom_point(aes(color = continent), size = .25) + 
    scale_x_log10() +
    geom_smooth(method = 'lm', color = 'black', linetype = 'dashed') +
    theme_bw()
```


> #### Challenge -- sizing points {.challenge}
>
> Modify the previous plot to size the points in proportion to the population of 
> the county.
>



### Multi-panel figures

Earlier we visualized the change in life expectancy over time across all
countries in one plot. 

```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, by = country)) +
    geom_line(aes(color = continent))
```

That's a mess! It's hard to see patterns because there is so much information on 
top of itself. We can split this out over multiple panels by adding another 
layer: **facet** panels. The `facet_wrap` layer takes a "formula" as its 
argument, denoted by the tilde (~). You can read this as "by continent". It 
tells R to draw a panel for each unique value of continent.

```{r facet}
ggplot(gapminder, aes(x = year, y = lifeExp, by = country)) +
    geom_line(aes(color = continent)) +
    facet_wrap(~ continent)
```



### Modifying text

To clean this figure up for a publication we need to change some of the text 
elements. The x-axis is way too cluttered, and the axis labels should read "Life 
expectancy" and "Year", rather than the column name in the data frame. Also, the 
country names don't fit in the strips at the top of each plot, and we might like 
to add an overall title to the figure.

We can do this by adding a few new layers. The `xlab` and `ylab` layers control 
the names of the x and y axes. The **theme** layer is where we change lots of 
minor details of a plot. here we'll rotate the axis text (with `angle`) so that 
they'll fit, and then make sure they're lined up right with `vjust` for 
vertical-adjustment. And let's add a plot title for fun; we can do that with 
`ggtitle` layer.

```{r cleaned up}
ggplot(gapminder, aes(x = year, y = lifeExp, by = country)) +
    geom_line(aes(color = continent)) +
    facet_wrap(~ continent) +
    xlab("Year") + 
    ylab("Life expectancy") + 
    theme_bw(base_size = 10) +
    scale_color_brewer(guide = "none", palette = "Dark2") +
    theme(axis.text.x = element_text(angle = 270, vjust = .5)) +
    ggtitle('Life expectancy over time by country')
```


### Saving plots

Now that we have that plot just the way we want it, let's write it to disk as an 
image. Of course, we could regenerate it by re-running our script, but we might 
want to make it available separately to show a collaborator or include in a 
document or presentation.

You can use the `ggsave` function to save a ggplot. By default, `ggave` will 
save the last generated plot in the same dimensions as your viewing pane. The 
only required argument to `ggsave` is the location of the file you want to write 
to. Using our project organization, let's write the last plot to 
`results/lifeExpByCountry.png`. `ggsave` will notices the extension of the 
filename and writes the file in that format. You can also save as .pdf, .jpg, 
and other file types.

```{r}
ggsave('results/lifeExpByYear.png')
```

Some of the country names are still getting squeezed, so let's make the canvas a 
little bigger, and we'll save a pdf file instead of a png. Let's also assign our 
plot to an object and explicitly save that plot to disk. You assign a ggplot 
object to a variable just like any other object.

```{r}
myplot <- 
  ggplot(gapminder, aes(x = year, y = lifeExp, by = country)) +
    geom_line(aes(color = continent)) +
    facet_wrap(~ continent) +
    xlab("Year") + 
    ylab("Life expectancy") + 
    theme_bw(base_size = 14) +
    scale_color_brewer(guide = "none", palette = "Dark2") +
    theme(axis.text.x = element_text(angle = 270, vjust = .5)) +
    ggtitle('Life expectancy over time by country')
ggsave(filename = 'results/lifeExpByYear-bigger.pdf', 
       plot = myplot, 
       width = 12, height = 8)
```


### Wrap-up

This is just a taste of what you can do with `ggplot2`. RStudio provides a
really useful [cheat sheet][cheat] of the different layers available, and more
extensive documentation is available on the [ggplot2 website][ggplot-doc].
Finally, if you have no idea how to change something, a quick Google search will
usually send you to a relevant question and answer on Stack Overflow with 
reusable code to modify!

[cheat]: http://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
[ggplot-doc]: http://docs.ggplot2.org/current/


> #### Challenge 5
>
> Create a density plot of GDP per capita, with a separate curve for each 
> continent, with color filled by continent. Save the plot to your `results/` 
> directory.
>
> **Advanced**:
> 
>   - Adjust the opacity (`alpha`) of the density curves so all are visible.
>   - Transform the-x axis to better visualize the data spread.
>   - Add a facet layer to panel the density plots by year.
>




## Challenge solutions

> #### Solution to challenge 1
>
> Modify the example so that the figure visualize how life expectancy has
> changed over time:
>
> ```{r ch1-sol}
> ggplot(gapminder, aes(x = year, y = lifeExp)) + geom_point()
> ```
>

> #### Solution to challenge 2
>
> In the previous examples and challenge we've used the `aes` function to tell
> the scatterplot **geom** about the **x** and **y** locations of each point.
> Another *aesthetic* property we can modify is the point *color*. Modify the
> code from the previous challenge to **color** the points by the "continent"
> column. What trends do you see in the data? Are they what you expected?
>
> ```{r ch2-sol}
> ggplot(gapminder, aes(x = year, y = lifeExp, color=continent)) +
>   geom_point()
> ```
>

> #### Solution to challenge 3
>
> Switch the order of the point and line layers from the previous example. What
> happened?
>
> ```{r ch3-sol}
> ggplot(gapminder, aes(x=year, y=lifeExp, by=country)) +
>  geom_point() + geom_line(aes(color=continent))
> ```
> 
> The lines now get drawn over the points!
>


> #### Solution to challenge 4
>
> Modify the color and size of the points on the point layer in the previous
> example.
>
> Hint: do not use the `aes` function.
>
> ```{r ch4-sol}
> ggplot(gapminder, aes(x = lifeExp, y = gdpPercap)) +
>  geom_point(size=3, color="orange") + scale_y_log10() +
>  geom_smooth(method="lm", size=1.5)
> ```
>

> #### Solution to challenge 5
>
> Create a density plot of GDP per capita, filled by continent.
>
> Advanced:
>  - Transform the x axis to better visualize the data spread.
>  - Add a facet layer to panel the density plots by year.
>
> ```{r ch5-sol}
> ggplot(gapminder, aes(x = gdpPercap, fill=continent)) +
>  geom_density(alpha=0.6) + facet_wrap( ~ year) + scale_x_log10()
> ```
>

# From the dplyr lesson

> #### Challenge -- Part 3 (Advanced)
> 
> The above plotting exercise asked you to plot summarized information, but it 
> is generally preferable to avoid summarizing before plotting. Can you generate 
> a plot that shows the information you calculated in Part 1 without 
> summarizing?  
>
> - Hint: `ggplot` interprets the `gapminder$year` as a numeric variable, which 
> may be okay, but there are some plot types for which you need `ggplot` to see 
> `gapminder$year` as a category. You can accomplish this by wrapping it in 
> `factor` -- e.g. `ggplot(gapminder, aes(x = factor(year) ...`
>

#### Resources

That is the core of `dplyr`'s functionality, but it does more. RStudio makes a 
great [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) 
that covers all the `dplyr` functions we just learned, plus what we will learn 
in the next lesson: keeping data tidy.


<!-- ## Challenge solutions -->

<!-- > #### Solution to challenge Subsetting {.challenge} -->
<!-- > -->
<!-- > - Subset the gapminder data to only Oceania countries post-1980. -->
<!-- > - Remove the continent column -->
<!-- > - Make a scatter plot of gdpPercap vs. population colored by country -->
<!-- > -->
<!-- > ```{r challenge subsetting a} -->
<!-- > oc1980 = filter(gapminder, continent == "Oceania" & year > 1980) -->
<!-- > oc1980less = select(oc1980, -continent) -->
<!-- > library('ggplot2') -->
<!-- > ggplot(oc1980less, aes(x = gdpPercap, y = lifeExp, color = country)) + -->
<!-- >   geom_point() -->
<!-- > ``` -->
<!-- > -->
<!-- > **Advanced** How would you determine the median population for the North  -->
<!-- > American countries between 1970 and 1980? -->
<!-- > -->
<!-- > ```{r challenge subsettingB} -->
<!-- > noAm = filter(gapminder, country == "United States" |  -->
<!-- >                 country == "Canada" | country == "Mexico" |  -->
<!-- >                 country == "Puerto Rico" & (year > 1970 & year < 1980) -->
<!-- >               ) -->
<!-- > noAmPop = select(noAm, pop) -->
<!-- > median(noAmPop) -->
<!-- > noAmPop -->
<!-- > as.integer(noAmPop) -->
<!-- > median(unlist(noAmPop)) -->
<!-- > ``` -->
<!-- > -->
<!-- > **Bonus** This can be done using base R's subsetting, but this class doesn't  -->
<!-- > teach how. Do the original challenge without the `filter` and `select`  -->
<!-- > functions. Feel free to consult Google, helpfiles, etc. to figure out how. -->
<!-- > -->
<!-- > ```{r challenge subsettingC} -->
<!-- > noAm2 = gapminder[(gapminder$country == "United States") | -->
<!-- >                     (gapminder$country == "Mexico") | -->
<!-- >                     (gapminder$country == "Canada") | -->
<!-- >                     (gapminder$country == "Puerto Rico") & -->
<!-- >                     ((gapminder$year > 1970) & -->
<!-- >                     (gapminder$year < 1980)),] -->
<!-- > median(noAm2$pop) -->
<!-- > ``` -->
<!-- > -->

<!-- > #### Solution to challenge MCQ: Data Reduction {.challenge} -->
<!-- > -->
<!-- > Produce a data.frame with only the names, years, and per-capita GDP of  -->
<!-- > countries where per capita gdp is less than a dollar a day sorted from most-  -->
<!-- > to least-recent. -->
<!-- > -->
<!-- > - Tip: The `gdpPercap` variable is annual gdp. You'll need to adjust. -->
<!-- > - Tip: For complex tasks, it often helps to use pencil and paper to  -->
<!-- > write/draw/map the various steps needed and how they fit together before  -->
<!-- > writing any code. -->
<!-- > -->
<!-- > What is the annual per-capita gdp, rounded to the nearest dollar, of the first  -->
<!-- > row in the data.frame? -->
<!-- > -->
<!-- > a. $278 -->
<!-- > b. $312 -->
<!-- > c. $331 -->
<!-- > d. $339 -->
<!-- >  -->
<!-- > ```{r data reduction A} -->
<!-- > dailyGDP = mutate(gapminder, onedayGDP = gdpPercap / 365) -->
<!-- > dailyGDP = filter(dailyGDP, onedayGDP < 1) -->
<!-- > dailyGDP = select(dailyGDP, country, year, gdpPercap) -->
<!-- > dailyGDP[1,] -->
<!-- > ``` -->
<!-- > -->
<!-- > **Advanced**: Use dplyr functions and ggplot to plot per-capita GDP versus -->
<!-- > population for North American countries after 1970. -->
<!-- > - Once you've made the graph, transform both axes to a log10 scale. There are -->
<!-- > two ways to do this, one by creating new columns in the data frame, and -->
<!-- > another using functions provided by ggplot to transform the axes. Implement -->
<!-- > both, in that order. Which do you prefer and why? -->
<!-- > -->
<!-- > ```{r data reduction B} -->
<!-- > noAm = filter(gapminder, country == "United States" |  -->
<!-- >                 country == "Canada" | country == "Mexico" |  -->
<!-- >                 country == "Puerto Rico" & year > 1970 -->
<!-- >               ) -->
<!-- > ggplot(noAm, aes(x = gdpPercap, y = pop, color = country)) + -->
<!-- >   geom_point() +  -->
<!-- >   scale_x_log10() + -->
<!-- >   scale_y_log10() -->
<!-- > # OR -->
<!-- > noAmlog10 = mutate(noAm, log10pop = log10(pop), -->
<!-- >                    log10gdp = log10(gdpPercap)) -->
<!-- > ggplot(noAmlog10, aes(x = log10gdp, y = log10pop, color = country)) + -->
<!-- >   geom_point() -->
<!-- > ``` -->
<!-- > -->

<!-- > #### Challenge: Data Reduction with Pipes -->
<!-- > -->
<!-- > Copy the code you (or the instructor) wrote to solve the previous MCQ Data  -->
<!-- > Reduction challenge. Rewrite it using pipes (i.e. no assignment and no nested  -->
<!-- > functions) -->
<!-- > -->
<!-- > ```{r data reduction pipes} -->
<!-- > # previous challenge with pipes -->
<!-- > dailyGDP = mutate(gapminder, onedayGDP = gdpPercap / 365) -->
<!-- > dailyGDP = filter(dailyGDP, onedayGDP < 1) -->
<!-- > dailyGDP = select(dailyGDP, country, year, gdpPercap) -->
<!-- > # BECOMES -->
<!-- > smallGDP = gapminder %>% -->
<!-- >   mutate(onedayGDP = gdpPercap / 365) %>% -->
<!-- >   filter(onedayGDP < 1) %>% -->
<!-- >   select(country, year, gdpPercap) -->
<!-- > smallGDP[1,] -->
<!-- > # OR, more fancy (without an intermediate temp variable) -->
<!-- > (gapminder %>% -->
<!-- >   mutate(onedayGDP = gdpPercap / 365) %>% -->
<!-- >   filter(onedayGDP < 1) %>% -->
<!-- >   select(country, year, gdpPercap))[1,] -->
<!-- > ``` -->
<!-- > -->

<!-- > #### Challenge -- Part 1  -->
<!-- > -->
<!-- > - Calculate a new column: the total GDP of each country in each year.  -->
<!-- > - Calculate the variance -- `var()` of countries' gdps in each year. -->
<!-- > - Is country-level GDP getting more or less equal over time? -->
<!-- > - Plot your findings. -->
<!-- >  -->
<!-- > ```{r challenge LessonEnd part1} -->
<!-- > varGDP = gapminder %>% -->
<!-- >   mutate(totalGDP = gdpPercap * pop) %>% -->
<!-- >   group_by(year) %>% -->
<!-- >   summarize(varTotGDP = var(totalGDP)) -->
<!-- > ggplot(varGDP, aes(x = year, y = varTotGDP)) + -->
<!-- >   geom_point() -->
<!-- > ``` -->
<!-- > -->

<!-- > #### Challenge -- Part 2 -->
<!-- > -->
<!-- > - Modify the code you just wrote to calculate the variance in both  -->
<!-- > country-level GDP and per-capita GDP.  -->
<!-- > - Do both measures support the conclusion you arrived at above? -->
<!-- >  -->
<!-- > ```{r challenge LessonEnd part2} -->
<!-- > varGDP = gapminder %>% -->
<!-- >   mutate(totalGDP = gdpPercap * pop) %>% -->
<!-- >   group_by(year) %>% -->
<!-- >   summarize(varTotGDP = var(totalGDP), -->
<!-- >             varPerCapGDP = var(gdpPercap) -->
<!-- >             ) -->
<!-- > ggplot(varGDP) + -->
<!-- >   geom_point(color = "red", aes(x = year, y = varTotGDP)) + -->
<!-- >   geom_point(color = "blue", aes(x = year, y = varPerCapGDP)) -->
<!-- > ``` -->

<!-- > #### Challenge -- Part 3 (Advanced)  -->
<!-- >  -->
<!-- > The above plotting exercise asked you to plot summarized information, but it  -->
<!-- > is generally preferable to avoid summarizing before plotting. Can you generate  -->
<!-- > a plot that shows the information you calculated in Part 1 without  -->
<!-- > summarizing? -->
<!-- > -->
<!-- > - Hint: `ggplot` interprets the `gapminder$year` as a numeric variable, which  -->
<!-- may be okay, but there are some plot types for which you need `ggplot` to see  -->
<!-- `gapminder$year` as a category. You can accomplish this by wrapping it in  -->
<!-- `factor` -- e.g. `ggplot(gapminder, aes(x = factor(year) ...` -->
<!-- > -->
<!-- > -->
<!-- > ```{r challenge LessonEnd part3, fig.width=8} -->
<!-- > gapminder %>% -->
<!-- >   mutate(totalGDP = gdpPercap * pop) %>% -->
<!-- > ggplot(aes(x = factor(year), y = totalGDP)) + -->
<!-- >   geom_violin() + -->
<!-- >   scale_y_log10()  -->
<!-- >  ``` -->
<!-- > -->

This lesson is adapted from the Software Carpentry: R for Reproducible 
Scientific Analysis [Creating Publication Quality Graphics materials](http://data-lessons.github.io/gapminder-R/04-ggplot.html)