---
title: "Data import"
---

> ## Learning objectives
>
> * Get comfortable importing different kinds of data
> * Understand the concept of "tidy data"

```{r, include = FALSE}
library(tidyverse)
```

## Reading csv data

Data come in many forms, and we need to be able to load them in R. For our own 
use and with others who use R, there are R-specific data structures we can use, 
like the .RDA file-type we just saw, but we need to be able to work with more 
general data types too. Comma-separated value (csv) tables are perhaps the most 
universal data structure. 

The species dataset provides genus and species information for animals caught
during a trapping survey. I downloaded it and put it in the data directory of my 
project. You will do the same in a minute.

We can read csv's with the `read.csv` function. The argument to `read.csv` is 
the location of the file, relative to your working directory. Since I saved the 
gapminder data to the `data` directory of my project, I can load it with this. 
Note the use of tab-completion to find the file and get it exactly right without 
typos. 

```{r, eval = FALSE}
read.csv('data/species.csv')
```

Whoa! What just happened? R executed the function and printed the result, just 
like when you enter `log(1)`. How do you store an object to a variable?

```{r}
species <- read.csv('data/species.csv')
```

Now, a data.frame called `species` is in my Environment, and I can see that it 
is a data.frame with `r nrow(species)` rows and `r ncol(species)` columns. 

> #### Challenge -- read csv data
>
> The species data are available at [this link](data/species.csv).  
> - Right click on the link to "save file as..."  
> - Save the .csv file in the `/data` directory of your project.  
> - Read the data with the `read.csv` function and assign it to the variable 
> `species`.  
> - Inspect the data.frame using whatever means you like
> - Use a dplyr verb to reorder the data frame in alphabetical order by genus
> 
> **Advanced challenge**
>
> Suppose you get a .csv file from a colleague in Europe. Because they use "," 
> (comma) as a decimal separator, they use ";" (semi-colons) to separate fields. 
> How can you read it into R? 
>
> Feel free to use the web and/or R's helpfiles.    
>

### Using `stringsAsFactors=FALSE`

By default, when building or importing a data frame, the columns that contain 
characters (i.e., text) are coerced (=converted) into the factor data type. 
Depending on what you want to do with the data, you may want to keep these 
columns as character. To do so, `read.csv()` and `read.table()` have an argument 
called `stringsAsFactors` which can be set to `FALSE`.

 > **Tip**: In most cases, it’s preferable to set `stringsAsFactors = FALSE` 
 > when importing your data, and converting as a factor only the columns that 
 > require this data type.

Compare the output of `str(surveys)` when setting `stringsAsFactors = TRUE` 
(default) and `stringsAsFactors = FALSE`:

```{r eval = FALSE, echo = TRUE}
surveys <- read.csv('data/combined.csv')

## Compare the difference between when the data are being read as
## `factor`, and when they are being read as `character`.
surveys <- read.csv("data/combined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/combined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)

```

## Reading messier data

Sometimes data can be a bit trickier to read because it isn't in tidy format.
If it is **close** to being tidy, we may be able add some more arguments to the
`read.csv()` call to to help R interpret how the file should be read. If there
are a few things that R will need help with, using the `read.table()` gives you
the most flexibility but you'll also have to always specify more information
about the file to help R interpret it.

Use [this link](data/wide_eg.csv) to download this dataset and inspect it with
a spreadsheet program. Why isn't it tidy?

Try using the `read.csv()` function on the file.

```{r, eval = FALSE}
read.csv("data/wide_eg.csv")
```

We need to tell R to skip 2 rows!

```{r, eval = FALSE}
read.csv("data/wide_eg.csv", skip = 2)
```

Or using `read.table()`...

```{r, eval = FALSE}
read.table("data/wide_eg.csv", header = TRUE, sep = ",", skip = 2)
```

## Reading directly from websites

We can use the `read.csv()` function to read files directly from a website if we
first call the `url()` function on the web address.

```{r wwwReadCsv, eval = FALSE}
read.csv(url("https://mikoontz.github.io/data-carpentry-week/data/wide_eg.csv"), skip = 2)
```

We can also use the `read_csv()` (note that's an underscore in the function
name, not a period) from the `readr` package which is part of the `tidyverse`.
In this case, we don't have to use the `url()` function.

```{r, eval = FALSE}
read_csv("https://mikoontz.github.io/data-carpentry-week/data/wide_eg.csv", skip = 2)
```

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new datasets to share
them with your collaborators or for archival.

Similar to the `read.csv()` function used for reading CSV files into R, there is
a `write.csv()` function that generates CSV files from data frames.

Before using `write.csv()`, we are going to create a new folder, `data_output`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data_output`
directory, so even if the files it contains are deleted, we can always
re-generate them.

<br>

This lesson is adapted from the Software Carpentry: R for Reproducible 
Scientific Analysis [Vectors and Data Frames materials](http://data-lessons.github.io/gapminder-R/03-data-types-subsetting.html)
and the Data Carpentry: R for Data Analysis and Visualization of Ecological Data 
[Exporting Data materials](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html).