---
title: "Using `map` Functions with `purrr`"
---

<br>

## R4DS: [the map functions](http://r4ds.had.co.nz/iteration.html#the-map-functions)

<br>

**Brandon's PRESENTATION**:

<iframe width="600" height="400" src="http://d-rug.github.io/images/20171026/20171023_DRUG_map_walk.html#1"></iframe>

 - [*Fullscreen*](http://d-rug.github.io/images/20171026/20171023_DRUG_map_walk.html#1)

**Brandon's SCRIPT: BELOW**

# Let's load tidyverse
``` {r setup}
library(tidyverse)
```

# R Data types
```{r vector intiation}
#initiate blank vectors of a specific type
#this can be useful when capturing data from for loops
a <- vector("logical", 5)
a
b <- vector("integer", 4)
b
c <- vector("double", 2) # can also use "numeric"
c
d <- vector("character", 10)
d
e <- vector("complex", 7)
e
f <- vector("raw", 3)
f
g <- list()
g
```
# NAs
```{r NA NA NA NA NA NA NA NA Batman}
#what do the NAs look like?
list(NA, NA_integer_, NA_real_, NA_character_, NA_complex_)
#not very exciting
```

# The NULL element

``` {r NULLs}
# NULL is the absence of a vector
a <- c(TRUE, FALSE, TRUE, FALSE, FALSE, NA) # logical vector 
a
a <- NULL
a

mtcars
mtcars[1] <- NULL #this gets rid of the mpg coloumn completely
mtcars

data(mtcars) # restore mtcars

```

# Assembling vectors into a list, dataframes and tibbles
```{r Vectors Victor, error = TRUE}
# let's put actual values into these
a <- c(TRUE, FALSE, TRUE, FALSE, FALSE, NA) #seems logical 
b <- c(1:7, NA_integer_) #Integers
c <- c(rnorm(8), NA_real_) #pull doubles from normal distribution
d <- c(LETTERS[1:3], NA_character_) #first 5 capital letters

e <- list(a, b, c, d) #make a list

e # check it twix

#Lists can contain anything of any length
#data.frame is a special type of list
# data.frame rows imply association so they need to line up

data.frame(a, b) # uh oh

a <- c(TRUE, FALSE, TRUE, FALSE, FALSE, NA, NA, NA) # make it 8 like b
df <- data.frame(a = a, b = b) # OH YEAH /Koolaidman.gif
df 
#tibbles are a more strict type of data.frame
# with
#     slightly more informative printing (not shown well here)
#     strict behavior around subsetting `[` -> tibble `[[` vector
#     does not partially match column names (this can get you into trouble)

as_tibble(df)

```
# Back to presentation

# Accessing elements inside vectors, lists, dataframes, and tibbles
``` {r list_df_tibble_subsetting, error = TRUE} 

#atomic vectors
a <- c(TRUE, FALSE, TRUE, FALSE, FALSE, NA) #logic vector 
a
a[1] 
a[[1]] # does the same thing
a[[[1]]] #lolwut R you doing?

#lists
list_example <- list(a = a, b= b, c = c, d = d, e = list(1, "A")) #make a named list
list_example
# single brackets returns a list with that element in it
# can reference by location or by name
list_example[1]
list_example["a"]

# double brackets return a vector 
list_example[[2]]
list_example[["b"]]

#stack [['s it to peel the onion back
list_example[["e"]] # remove one layer
list_example[["e"]][[2]] # remove the second layer

#dolla dolla bills y'all (works like [[, with named vectors)
list_example$c

#same for dataframes
mtcars[1]
mtcars[[1]]

# and for tibbles
tbbl <- tibble(x = 1:3, y = LETTERS[1:3])
tbbl[1]
tbbl["x"]

tbbl[[2]]
tbbl[["y"]]
```


# Back to presentation


# When to make a function in R
``` {r Y_make_a_function}
#Simple dataframe
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
# What are we doing here?
df$a <- (df$a - min(df$a, na.rm = TRUE)) /  
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) /  
  (max(df$b, na.rm = TRUE) - min(df$b, na.rm = TRUE))
df$c <- (df$c - min(df$a, na.rm = TRUE)) /  
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) /  
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

```


``` {r function01, error = TRUE, eval = FALSE}
df$d <- (df$d - min(df$d, na.rm = TRUE)) /  
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```


``` {r function02, error = TRUE, eval = FALSE}
# pull out what's common
x <- (x - min(x, na.rm = TRUE)) /  
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```


``` {r function03, error = TRUE, eval = FALSE}

#simplify more and refactor if that makes sense
x <- (x - min(x, na.rm = TRUE)) /  
  diff(range(x, na.rm = TRUE))
```

``` {r function04, error = TRUE, eval = FALSE}
# rewrite as a function
rescale_0_1 <- function(x) {
  (x - min(x, na.rm = TRUE)) /  
  diff(range(x, na.rm = TRUE))
}
```

``` {r test function}
rescale_0_1 <- function(x) {
  (x - min(x, na.rm = TRUE)) /  
  diff(range(x, na.rm = TRUE))
}

testvec <- 1:11
testvec
rescale_0_1(testvec)
```

``` {r use_function}
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
rescale_0_1 <- function(x) {
  (x - min(x, na.rm = TRUE)) /  
  diff(range(x, na.rm = TRUE))
}

out <- df
out$a <- rescale_0_1(df$a)
out$b <- rescale_0_1(df$b)
out$c <- rescale_0_1(df$c)
out$d <- rescale_0_1(df$d)
out
```

# Reduce iteration with a for loop
``` {r forloops}
out <- df #make a copy to store scaled data
for (i in 1:ncol(df)) { # loop through each element
  out[i] <- rescale_0_1(df[[i]]) #apply function and store it to out
}
out
```

# seq_along

``` {r for_seq_along_loops}
out <- df #make a copy to store scaled data
for (i in seq_along(df)) {
  out[i] <- rescale_0_1(df[[i]])
}
out
```

# map to iterate
```{r WHAT_IS_THIS_MAPGIC}
map(df, function(x) rescale_0_1(x)) #map over the columns in df
```
# let's look at another more simple example
# This list contains two integer vectors and we want the mean of each one
``` {r simplemap}
l = list(a=1:10, b = 10:100)
map(l, function(x) mean(x, na.rm = TRUE))
```

# We know we want one double back from each list
# specify and you get a vector back instead of a list
``` {r rightmap}
l = list(a=1:10, b = 10:100)
map_dbl(l, function(x) mean(x, na.rm = TRUE))
```

# You can coerce these to character so this works
#   but I don't think this is very helpful
``` {r okishmap}
map_chr(l, function(x) mean(x, na.rm = TRUE))
```


# These are type-strict functions though
# get it wrong and you'll know
``` {r wrongmap, error = TRUE}
l = list(a=1:10, b = 10:100)
map_lgl(l, function(x) mean(x, na.rm = TRUE))
```

# Map Examples
# Map use case #1: Read in files
Based upon this tutorial
http://data.library.virginia.edu/getting-started-with-the-purrr-package-in-r/

``` {r simple_read_csvs, message = FALSE}
#this will work if you clone the github repository
fileloc <- file.path(getwd(), "stocks")
#create a list of files to read in that end in csv
files <- list.files(fileloc, pattern = "csv$", full.names = TRUE) 

#this reads in each csv independently into a list
dat2 <- map(files, read_csv)
dat2

```

# Let's make it a function
``` {r make_it_functional, message = FALSE}
fileloc <- file.path(getwd(), "stocks")

read_folder <- function(folder) {
  # get files that end with csv or CSV
  files <- list.files(folder, pattern = "(csv|CSV)$", full.names = TRUE) 
  #this reads in each csv independently into a list
  map(files, read_csv) # last thing is returned
}

read_folder(fileloc)
```


# What if you want to change the type of file you're reading?
``` {r fancify, message = FALSE}
fileloc <- file.path(getwd(), "stocks")

read_folder <- function(folder, pattern, read_fun) {
  # get files that end with csv or CSV
  files <- list.files(folder, pattern = pattern, full.names = TRUE) 
  #this reads in each csv independently into a list
  map(files, read_fun)
}

read_folder(folder = fileloc, pattern = "(csv|CSV)$", read_fun = read_csv)
# OR more succinctly
# read_folder(fileloc, "(csv|CSV)$", read_csv)
```

# We've got 3 independent datasets, this isn't always helpful
# let's combine them into one, but we need to label them by their filenames

``` {r combinify, message = FALSE}
fileloc <- file.path(getwd(), "stocks")

# This function reads a file and appends a column called "name" for ID
read_fun <- function(file, FUN) {
		dat <- FUN(file) # read in file
		dat$name <- tools::file_path_sans_ext(basename(file)) #record the filename for later grouping
		dat #bring data back
  }

# This function reads a folder for files of given pattern
# then applies the given read function to read them in read_fun
read_files_bind <- 	function(folder, pattern, FUN) {
    # get files that end with csv or CSV
    files <- list.files(folder, pattern = pattern, full.names = TRUE) 
    map_dfr(files, read_fun, FUN)
  }

# Read everything in at once and bind it all together
read_files_bind(folder = fileloc, pattern = "(csv|CSV)$", FUN = read_csv)
```

# List columns
# tidyr::nest() embeds tibbles in a column
``` {r nesting}
cars_nested <-
  mtcars %>%            # motortrend car dataset
    group_by(cyl) %>%   # group by cylinder
    nest(.key = "data") # nest data by cyl, default key is "data"

cars_nested
```  
  
# This is neat, but how to we interact with the nested data?
# map!
## let's make a linear model for each subset, pull out the coeffiecients with broom, and plot them
``` {r model_extracts}
library(broom)

mtc_modeled <- 
  mtcars %>%            # motortrend car dataset
    group_by(cyl) %>%   # group by cylinder
    nest(.key = "data") %>%         # nest data by cyl, default key is "data"
    mutate(model = map(data, ~ lm(mpg ~ disp, data = .x))) %>%  # make models
    mutate(coeff = map(model, tidy)) %>%                        #pull out coefficients
    mutate(slope = map_dbl(coeff, ~ pull(.x, estimate)[2])) %>% # pull slope from coefficients
    mutate(intercept = map_dbl(coeff, ~ pull(.x, estimate)[1])) # pull intercept from coefficients
  
 
ggplot() +
  # plot the original data 
  geom_point(data = mtcars, aes(x = disp, y = mpg, color = as.factor(cyl))) +
  # plot the models
  geom_abline(data = mtc_modeled, aes(slope = slope, intercept = intercept, color = as.factor(cyl)))
  
  
```


#Back to presentation

# Iterating over more than one list with map2 and pmap
``` {r map2_pmap_usage example}
n =  list(5,10,20)
mu = list(50,30,10)
sd = list(10,5,1)

# how does it look as we go from map to map2 to pmap?
map(n, rnorm)
map2(n, mu, rnorm)
pmap(list(n, mu, sd), rnorm)

#technically, pmap can do it all
pmap(list(n), rnorm)
pmap(list(n, mu), rnorm)

# formula interface
map(n, ~rnorm(.))
map2(n, mu, ~ rnorm(.x, .y))
pmap(list(n, mu, sd), ~rnorm(..1, ..2, ..3)) # ..4 and so on for as many as you need

# CAUTION: position matters with map2 & pmap
map2(n, mu, rnorm)
map2(mu, n, rnorm)
pmap(list(n, mu, sd), rnorm)
pmap(list(mu, n, sd), rnorm)

# if you need non-default ordering, specify with full function call
map2(mu, n, function(x, y) rnorm(y, x))
# or using the formula interface
map2(mu, n, ~ rnorm(.y, .x))

# pmap you can name your list to get around positional ordering
pmap(list(mean=mu, n=n, sd=sd), rnorm)
# OR you can specify with a full function call
pmap(list(mu, n, sd), function(mu, n, sd) rnorm(n, mu, sd))
# OR with the formula interface
pmap(list(mu, n, sd), ~ rnorm(..2, ..1, ..3))
```

#Back to Presentation


# Walk/walk2/pwalk are used to output side-effects
``` {r simple walk}
asdf <- list(a = LETTERS[1:10], b = 1:15, c = FALSE)
# we just want to print each element of this list
walk(asdf, print)
```

# Let's make some plots and plot them
``` {r walking_to_plotting}
# make a simple plotting function
plot_points <- function(t) ggplot(t, aes(mpg, wt)) + geom_point()

mtcars %>%                 # motortrend car dataset
  group_by(cyl) %>%       # group by cylinder
  nest(.key = "data") %>% # nest by cylinder
  mutate(plot = map(data, plot_points)) %>% # we made a graph for each subset
  pull(plot) %>% #pull the plots out as a list
  walk(., print) # print them
```

# Nice, but I wanted to save them; walk2/pwalk

``` {r walk2_saving_plots}
# make a simple plotting function
plot_points <- function(t) ggplot(t, aes(mpg, wt)) + geom_point()

mtc_plot_n_save <-
  mtcars %>%                 # motortrend car dataset
    group_by(cyl) %>%       # group by cylinder
    nest(.key = "data") %>% # nest by cylinder
    mutate(plot = map(data, plot_points)) %>% # we made a graph for each subset
    mutate(filepath = map_chr(cyl, ~ file.path(getwd(), paste0(.x, ".png")))) #make the save paths


walk2(mtc_plot_n_save$plot, mtc_plot_n_save$filepath, ~ ggsave(.y, .x)) # save them out
```



``` {r pwalk_output_all_the_things}

# make a simple plotting function
plot_points <- function(t) ggplot(t, aes(mpg, wt)) + geom_point()

#function that prints cylinder info, ggplot, and table 
print_func <-
  function(info, plot, table) {
    print(paste0("This is for ", info, " cylinders.\n"))
    print(plot)
    print(table)
  }

mtc_plot_n_save <-
  mtcars %>%                 # motortrend car dataset
    group_by(cyl) %>%       # group by cylinder
    nest(.key = "data") %>% # nest by cylinder
    mutate(plot = map(data, plot_points)) %>% # we made a graph for each subset
    mutate(filepath = map_chr(cyl, ~ file.path(getwd(), paste0(.x, ".png")))) #make the save paths

# print all the info
pwalk(list(info = mtc_plot_n_save$cyl, plot = mtc_plot_n_save$plot, table = mtc_plot_n_save$data), print_func) 
```

# Back to Presentation
