---
title: "Untitled"
author: "Martha Wohlfeil"
date: "1/3/2020"
output: html_document
---
## Example 2: Pressure Transducer (Solinst) Data Logger 

Let's practice with another data file, a logger that records water temperature and river stage every 15 minutes from the NF American River. You can download the file by right clicking on the [webpage](../data/2015_NFA_solinst_08_05.csv), or simply use the web URL. We've practiced both now. (The file is also available on the R-DAVIS website in the [Resources->Datasets](https://gge-ucd.github.io/R-DAVIS/resources.html) tab.)

*This dataset has 35,038 observations for water temperature, and stage (pressure transducer info).*

 > Task 1: Read in and assign the data to your environment using either the direct webpage URL, or the downloaded file.

```{r solinst_data, eval=T}

# pressure transducer data

# read in file directly from a weblink
dat <- read.csv("https://github.com/gge-ucd/R-DAVIS/raw/master/data/2015_NFA_solinst_08_05.csv", skip=13)

library(lubridate, warn.conflicts = F)
library(dplyr, warn.conflicts = F)

summary(dat)
names(dat)

```

 > ### Task 2
 - Make a `datetime` column and then...
 > - Calculate Weekly **Mean, Max, Min** Water Temperatures and plot as point or line plot *(HINT: `dplyr` function)*
 > - Calculate Hourly mean **Level** for April through June and make a line plot (using `ggplot2`). 
 > - Make a ggplot geom_pointrange plot of the avg. max, min monthly temperatures.

**Solutions**

```{r solution2, eval=T, echo=F}

# make datetime column
dat$datetime <- paste0(dat$Date, " ", dat$Time)

# format as POSIXct using lubridate
dat$datetime <- ymd_hms(dat$datetime)

library(ggplot2)

dat %>% 
  mutate(yr = year(datetime),
         mon = month(datetime),
         wk = week(datetime),
         DOY = yday(datetime),
         hr = hour(datetime))

head(dat)

# Weekly Plot
dat %>%  select(Level, Temperature, wk) %>% # select only cols we want
  group_by(wk) %>% 
  summarise_each(funs(mean, max, min)) %>%  # apply functions to each column
  ggplot() + geom_point(aes(x=wk, y=Temperature_max), color="gray40", fill="maroon", pch=21) +
  geom_line(aes(x=wk, y=Temperature_mean), color="red3")+
  geom_point(aes(x=wk, y=Temperature_min), color="gray40", fill="navyblue", pch=21) + xlab("July 2015: Week")


# Hourly July Plot
dat %>%  filter(mon>=4 & mon<=6) %>% 
  select(Level, DOY, hr, yr) %>%
  group_by(yr, DOY, hr) %>% 
  summarise(mean_Level=mean(Level)) %>% 
  mutate(datetimes = ymd_hms(strptime(paste0(yr,"-",DOY," ",hr, ":00"), format="%Y-%j %H:%M"))) %>% 
  ggplot() + geom_line(aes(x=datetimes, y=mean_Level), color="forestgreen") + xlab("Date") + ylab("Hourly Stage (m)")

# Monthly Plot
dat %>%  
  select(Temperature, mon) %>% # select only cols we want
  group_by(mon) %>% 
  summarise_each(funs(mean, max, min)) %>%  # apply functions to each column
  ggplot() + geom_pointrange(aes(x=mon, y=mean, ymax=max, ymin=min), fill="maroon", size=1, pch=21) + xlab("Avg Monthly Water Temperature: Months")

```