---
title: "Data Import and Export"
---

<br>
<div class = "blue">
## Learning objectives

* Get comfortable importing different kinds of data
* Understand the concept of "tidy data"
</div>

<br>

```{r, include = FALSE}
library(tidyverse)
```

## Reading csv data

Data come in many forms, and we need to be able to load them in R. For our own 
use and with others who use R, there are R-specific data structures we can use, such as the `.rda` (`.RDATA`), or the `.rds` file-types. We'll talk about those in more detail later. However, we need to be able to work with more general data types too. Comma-separated value (`.csv`) tables are perhaps the most universal data structure. 

This [species dataset](https://gge-ucd.github.io/R-DAVIS/data/species.csv) provides genus and species information for animals caught
during a trapping survey. I downloaded it and put it in the `data` directory of my 
project. Click on the link and download the file.

## Review: What's the main difference between `read.csv` and `read_csv`? 

#### Answer: `stringsAsFactors=FALSE`

By default, when building or importing a data frame, the columns that contain 
characters (i.e., text) are coerced (=converted) into the factor data type when we use the `read.csv`. 
Depending on what you want to do with the data, you may want to keep these 
columns as character. To do so, `read.csv()` and `read.table()` have an argument 
called `stringsAsFactors` which can be set to `FALSE`.

When we use the `tidyverse` `read_csv` all character columns are automatically kept as characters, NOT coherced into factors. 

Let's look at the same data file read in with `read.csv` and `read_csv`
```{r}
surveys_dot <- read.csv('data/combined.csv')

surveys_underscore <- read_csv("data/combined.csv")

#using stringAsfactor

surveys_dot2 <- read.csv("data/combined.csv", stringsAsFactors = FALSE)

#can use as.character or as.factor to change single column types 
```

<div class = "blue">
### Advanced challenge

Suppose you get a .csv file from a colleague in Europe. Because they use `,` (comma) as a decimal separator, they use `;` (semi-colons) to separate fields.  
 - How can you read it into R? Feel free to use the web and/or R's helpfiles.  
</div>

<br>

## Reading messier data

Sometimes data can be a bit trickier to read because it isn't in tidy format.
If it is **close** to being tidy, we may be able add some more arguments to the
`read_csv()` call to to help R interpret how the file should be read. 

Use [this link](data/wide_eg.csv) to download this "wide" dataset and inspect it with
a spreadsheet program. Why isn't it tidy?

Try using the `read.csv()` function on the file.

```{r, eval = FALSE}
read_csv("data/wide_eg.csv")
```

We need to tell R to skip 2 rows!

```{r, eval = FALSE}
read_csv("data/wide_eg.csv", skip = 2)
```


We can use the `read_csv()` function to read files directly from a website. 

```{r, eval = FALSE}
read_csv("https://mikoontz.github.io/data-carpentry-week/data/wide_eg.csv", skip = 2)
```

## Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new datasets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

Before using `write_csv()`, we are going to create a new folder, `data_output`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data_output`
directory, so even if the files it contains are deleted, we can always
re-generate them.

 - Type in `write_` and hit TAB. Scroll down and take a look at the many options that exist for writing out data in R. Use the F1 key on any of these options to read more about how to use it.

## Using `save` and `load`

We've seen how to save out specific data types in R. One additional option is you can save anything in your workspace or working "environment". You can save a single object, or multiple objects together. This is best done using the `.rda` file type. The **.Rda** file type is a great way to save objects from R that you have already tidied or modeled, and you want to maintain the structure and format of the data. Because these are native "R" file types, they can be read back in and will appear exactly as when you saved them. 

Let's assume we have the "surveys_underscore" object and the "surveys_dot2" objects in our environment, and we want save them both in one file. We'll use `save()` and the `.rda` filetype.

```{r, eval=FALSE}
save(surveys_underscore, surveys_dot2, file = "data_output/surveys_types.rda")
```

To re-load these back into R, we use the `load` function. 

```{r loaddata, eval=F}

load("data_output/surveys_types.rda")

```

### `.rds` vs. `.rda`

Why use one vs. the other? What do these file types provide that a simple csv doesn't provide? 

Short answer is they maintain not only the structure, but also the format and data types within your data sets. However it appears in your environment in R is exactly how it will be saved (with `save()`), and then read back in (with `load()` or `readRDS`. This can save time and code, once you have your data in a format/shape you are happy with. An added bonus is the format (`rda` and `rds`) are both highly compressed, and can save significant space on your hard drive. 

The main differences between the two:

 - `.rda` allows saving multiple objects together in one file, or one single object
 - `.rds` can only save file/object
 
For example, let's use the surveys dataset we loaded earlier and save a few `rda` and `rds` files.
 
```{r rda_vs_rds, eval=F, echo=T}

library(dplyr)

# load data
surveys <- read_csv('data/combined.csv') # the combined.csv is 3.1 MB in size

# filter to years > 2000 and only rodents
rodents2001 <- surveys %>% filter(year > 2000 & taxa == "Rodent")

# filter to only birds
birds <- surveys %>% filter(taxa=="Bird")

# RDA files: now we can save all these objects together:
save(surveys, rodents2001, birds, file = "data/example.rda") # this file is 245 kb (0.2 MB)

# RDS files: this can only be one file:
saveRDS(rodents2001, file = "data/rodents2001.rds")

# to load an rds file back in:
any_name_i_choose <- readRDS(file = "data/rodents2001.rds")

# but won't work with rda
nope <- readRDS(file = "data/example.rda")

# same with trying to assign a name to an rda file on load
doublenope <- load(file = "data/example.rda") # wait ...check out "doublenope"

doublenope # created a character vector of the objects...but didn't assign any to the name nope

# [1] "surveys"     "rodents2001" "birds"   
``` 

One extremely time-saving use of `.rds` files is when you've fit a large statistical model that gets saved as a model object. For example, the `brms` and `lme4` packages used for multi-level models will return large model objects that contain lots of information about the model. Some models can take hours, days, even **weeks** to finish fitting, so it can be useful to save a fully fitted model object as a `.rds` file. You can then load that model object back into R and work with it however you want.

## Saving Figures and Plots 

A plot you created with ggplot or another plotting package can be saved as .JPEGS (or .tiff, .img, etc) onto you. For any ggplot objects, we reccomend using `ggsave`. 

Let's start by making a plot together with our `survey` data using only data that is for Rodents in years after 2000:  

```{r, warning = F, message = F}

# load data
surveys <- read_csv('data/combined.csv') # the combined.csv is 3.1 MB in size

# filter to years > 2000 and only rodents
rodents2001 <- surveys %>% filter(year > 2000 & taxa == "Rodent")

rodents2001 %>% ggplot()+
        geom_point(aes(x=hindfoot_length, y = weight, color = sex))
```
`ggsave` allows us to easily save this plot. First, let's create a new folder in this project called `figures`. Let's save all the figures we create to that folder. `ggsave` will default to saving the last plot you created, however, we think it is always a good idea to specify exactly which plot you want saved. To do that, we have to save our plot as an object. 

```{r, eval = F}
#saving plot as an object
rodent_plot <- rodents2001 %>% ggplot()+
        geom_point(aes(x=hindfoot_length, y = weight, color = sex))

ggsave("figures/rodent1.png", rodent_plot)
```
#### With `ggsave` you can save images as

 - .png, .jpeg, .tiff, .pdf, .bmp, or .svg

#### Other arguments of `ggsave`

 - `scale` can scale the image (multiplicative scaling factor)
 - `width` and `height` let you specify the size of the image in `units` that you specify
 - `dpi` can change the quality of the image; for publication graphs we suggest over 700 dpi

```{r, eval = F}

ggsave("figures/rodent2.png", rodent_plot, width = 4, height = 3, units = "in", dpi = 500)

```


## Excel & GoogleSheets

 - [`readxl`](http://readxl.tidyverse.org/) (Part of `tidyverse`)
 - Jenny Bryan's `googlesheets` or [`googledrive`](http://googledrive.tidyverse.org/) packages

## Other File Types

 - Using the `foreign` package
 - reading in `.dbf`, *Stata*, *SAS*, *SPSS*, `.shp`, `.netcdf`, `raster`, `.kml`, `gpx`, `xml`, `geojson`, `json`, etc....
 
## `rio` Package

One other data import/export package to check out is called `rio`. It uses two main functions: `import` and `export`, and it automatically detects the file extension you're using, then picks the fastest function out of a bunch of different specialized packages. It can streamline your data import/export and give you more consistent data frames when you're working with lots of different file types.

```{r}
#install.packages("rio")
library(rio)
export(mtcars, "data/mtcars.csv")
export(mtcars, "data/mtcars.rds")
export(mtcars, "data/mtcars.tsv")
```

You can read more about `rio` on its [Github page](https://github.com/leeper/rio).

<br>

This lesson is adapted from the Software Carpentry: R for Reproducible 
Scientific Analysis [Vectors and Data Frames materials](http://data-lessons.github.io/gapminder-R/03-data-types-subsetting.html)
and the Data Carpentry: R for Data Analysis and Visualization of Ecological Data 
[Exporting Data materials](http://www.datacarpentry.org/R-ecology-lesson/03-dplyr.html).