---
title: "date_time_extras"
author: "Martha Wohlfeil"
date: "12/26/2019"
output: html_document
---

## Background Reading

These resources are very useful. Please read through them.

 - This is a great overview of times and dates in R from [Bonnie Dixon's talk at the Davis R-Users Group](http://www.noamross.net/blog/2014/2/10/using-times-and-dates-in-r---presentation-code.html)

 - [A very comprehensive and detailed look at using dates in R](https://www.stat.berkeley.edu/~s133/dates.html)

 - [The vignette on using lubridate](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)
 
 
 
### Date Calculations

It's possible to use dates in calculations as well, as long as they are formatted in the same data type (i.e., both are `date` in this case).

**Make some fake dates**

```{r fakedates, eval=T, echo=T}

# format a value as `date`
dt1 <- as.Date("2017-07-11")
dt1

# non-standard formats must be specified using format=
dt2 <- as.Date("04/22/2016", format = "%m/%d/%Y")
dt2

dt3 <- as.Date("October 6, 2014", format = "%B %d, %Y")
dt3

```

**Now some calculations**

```{r datecalc1}
# subtract differences in dates
print(dt1 - dt2)
```

```{r datecalc2}
# or look at differences in alternate scales
print(difftime(dt1, dt2, units = "weeks"))

```

You can also sequence along by a given interval to create a timeseries. 

```{r}
six.weeks <- seq(dt1, length = 6, by = "week")
six.weeks
```

 > ### Challenge:
 > Create a sequence of 10 dates with a 2 week interval starting on 2011-02-15
 
```{r challenge2, echo=F, eval=F}
startingDate <- as.Date("2011-02-15")
tendates <- seq(startingDate, length=10, by=14)
tendates
```

## Working with Timezones & Datetimes

Similar to `as.Date` there are two functions for each datetime class in R. `as.POSIXct` and `as.POSIXlt`. The arguments are the largely the same, but two important components to consider are a `strptime` function to help format the datetime (similar to `as.Date`), and a timezone argument `tz`, for adding the appropriate timezone.

Keep in mind the default datetime format is `YMD HM` or `YMD HMS`, so if you use that format you won't need to use the `strptime` option.

### **as.POSIXct**

 - Example 1 (default doesn't need `strptime`): `as.POSIXct("2016-01-01 12:00", "America/Los_Angeles")` = `r as.POSIXct("2016-01-01 12:00", "America/Los_Angeles")`
 - Example 2 (24 hr time): `as.POSIXct(strptime("2016/04/05 14:47",format = "%Y/%m/%d %H:%M"), "America/Los_Angeles")` = `r as.POSIXct(strptime("2016/04/05 14:47",format = "%Y/%m/%d %H:%M"), "America/Los_Angeles")`
 - Example 3 (12 hr time): `as.POSIXct(strptime("2016/04/05 4:47 PM",format = "%Y/%m/%d %H:%M PM"), "America/Los_Angeles")` = `r as.POSIXct(strptime("2016/04/05 4:47 PM",format = "%Y/%m/%d %H:%M PM"), "America/Los_Angeles")`
 
 ## Datetimes without Timezones (and `chron`)

The `chron` package may be helpful for these tasks, however, this may also be a suitable use of the POSIXlt class.

 - `chron::as.chron("2013-07-24 23:55:26")` = `r chron::as.chron("2013-07-24 23:55:26")`
 - `chron::as.chron("07/25/13 08:32:07", "%m/%d/%y %H:%M:%S")` = `r chron::as.chron("07/25/13 08:32:07", "%m/%d/%y %H:%M:%S")`
 

 

```{r}


# or with custom legend
# use viridis color palette 
library(viridis)
#viridis(4) # "#440154FF" "#21908CFF" "#FDE725FF"
cols1 <- c("Temp-2m"="#440154FF", "Temp-10m" = "#35B779FF", "Temp-Towertop" = "#31688EFF")

df %>% 
  group_by(mon) %>% 
  summarize(avg_temp_2m = mean(temp_C_2m),
            avg_temp_10m = mean(temp_C_10m),
            avg_temp_tower = mean(temp_C_towertop)) %>% 
  ggplot() + geom_point(aes(x=mon, y=avg_temp_2m, fill="Temp-2m"), size=5, pch=21)+
  geom_point(aes(x=mon, y=avg_temp_10m, fill="Temp-10m"), size=5, pch=21) + 
  geom_point(aes(x=mon, y=avg_temp_tower, fill="Temp-Towertop"), size=5, pch=21) +
  ylab("2001 Avg Monthly Temperature (C)") + xlab("Month") + 
  scale_fill_manual(name="Temperature (C)",values=cols1) +
  theme(legend.position = "bottom")
 
```
